### 2012-05-18 final_paper/final_paper.tex
added the final paper tex

\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
\author{
        Russell Cohen \\
            \and
        Thomas Georgiou \\
            \and
        Pedram Razavi
}
\date{\today}

\documentclass[12pt]{article}

\begin{document}
\maketitle

\begin{abstract}
This is the paper's abstract \ldots
\end{abstract}

\section{Introduction}
In this project we aim to analyze and improve the performance of Python’s dictionary or dict. 
\paragraph{Outline}

\paragraph{Motivation}
Python heavily utilizes dictionary data structure for several of its internal language features. This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the the user’s program code does not explicitly use a dictionary. For instance, a dictionary can be instantiated to pass keyword arguments to a function. This means a dictionary could potentially be created and destroyed on every function call.  Some other examples of internal usage of dictionaries in Python are as follows:

Class method lookup
Instance attribute lookup and global variables lookup
Python built-ins module
Membership testing

Later in this paper, we compare the performance of our new dictionary implementation for some of these special cases. The aforementioned applications of the dictionaries further highlights the need for the fast instantiation and deletion of dictionaries so that less memory is utilized once running a program. Aside from the issue of memory usage, to further enhance the runtime of a program we need fast key/value storage and retrievals which is the main focus of this paper. Therefore it becomes clear that a better dictionary structure will have significant effects on total Python performance.

\paragraph{Current Python Implementation
}

In this section we give a brief overview of the key implementation details of dictionaries in CPython 2.7.3.
We should note that Python’s dictionary supports several different data types as keys. However as we noted earlier, the dictionaries underlying class instances and modules have only strings as keys. Therefore if we can

 but contains a specialized string-key-only version for performance in this common case. In this project we will focus more on the string data type as the key since most of the mentioned language features use only strings. The current Python dictionary implementation uses an iterative XORed polynomial terms for calculating the hash of the strings. The collision resolution strategy currently used is open addressing with a custom non-linear probing scheme. We will try to implement a hopefully more performant alternative to this strategy consisting of variants of simple tabulation hashing for variable length strings with linear probing.

The Theoretical Results and Way to improve implementation
Recently, Patrascu et al. [2] showed that simple tabulation hashing provides unexpectedly strong guarantees and is constant competitive with a truly random hash function for hashing with linear probing. They also showed some experimental evaluations which proved that simple tabulation on fixed input length can be fast and robust when used with linear probing and cuckoo hashing.   We plan to implement this scheme and analyze how it performs as a replacement for Python’s dict with the hopes of yielding better performance.

Technical details of implementation
Main Results
Benchmarks on real world apps
Conclusion and future work

\section{Previous work}\label{previous work}
A much longer \LaTeXe{} example was written by Gil~\cite{Gil:02}.

\section{Results}\label{results}
In this section we describe the results.

\section{Conclusions}\label{conclusions}
We worked hard, and achieved very little.

\bibliographystyle{abbrv}
\bibliography{main}

\end{document}

### 2012-05-19 docs/paper.tex
Commit paper into latex

\documentclass[11pt]{article}
\title{Improving Python Hashtable Performance with Simple Tabulation and Linear  Probing}
\author{Russell Cohen, Thomas Georgiou, Pedram Razavi}
\begin{document}
\maketitle
\section{Abstract} %or is there a \abstract?
Here we implement simple tabulation hashing with linear probing in Python dictionary based on the recent theoretical result of Patrascu et al. Although we achieved small performance wins on the order of 4%, we were unable to achieve significant performance wins.  The time and memory required to compute the ST hash are too large to justify the speedup garnered by switching to linear probing.

\section{Introduction}
In this project we aim to analyze and improve the performance of Python’s dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
1. Clean, elegant near-pseudo code syntax leads to clear code 
2. Widely distributed and available
3. Large quantity of external libraries for specific features

Recently, we the advent of packages like numPy, sciPy and pyPy python has began to also be used (with questionable merit) as a performance computing language due to its ease of use.

However, because Python is an interpreted language it is difficult to operate on a similar performance level to compiled languages due to the sheer overhead of the interpreter.

In order to attempt to improve the performance of the python language as a whole we plan to attempt to improve the performance of python dictionaries.

\subsection{Motivation}
Python heavily utilizes dictionary data structure for several of its internal language features. This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the the user’s program code does not explicitly use a dictionary. 
Take for example, the following code which we ran in the interpreter:

\begin{verbatim}
>> i = 2 
>> b = i * 2
>> b
10
\end{verbatim}

One of the features that we added to the python source was a conditional flag that allowed us view dictionary statistics (collisions, probes, accesses, etc.) for a given invocation of the Python binary.  With that modification we were able to see that this code used 92 lookups in string dictionaries and 24 lookups in general purpose dictionaries for a total of 116 dictionary hits. 116 hits in a 3 line piece of code which never uses dictionaries explicitly is quite significant.  We explain this extensive dictionary usage below.

For instance, a dictionary can be instantiated to pass keyword arguments to a function. This means a dictionary could potentially be created and destroyed on every function call.  Some other examples of internal usage of dictionaries in Python are as follows:
1. Class method lookup
2. Instance attribute lookup and global variables lookup
3. Python built-ins module
4. Membership testing

Later in this paper, we compare the performance of our new dictionary implementation for some of these special cases. The aforementioned applications of the dictionaries further highlights the need for the fast instantiation and deletion of dictionaries so that less memory is utilized once running a program. Aside from the issue of memory usage, to further enhance the runtime of a program we need fast key/value storage and retrievals which is the main focus of this paper. Therefore it becomes clear that a better dictionary structure will have significant effects on total Python performance.

\section{Current Python Implementation}

In this section we give a brief overview of the key implementation details of dictionaries in CPython 2.7.3.
We should note that Python’s dictionary supports several different data types as keys. However as we noted earlier, the dictionaries underlying class instances and modules have only strings as keys. Therefore optimizing a dictionary which have string-only keys is crucial to the runtime of any Python program. CPython accommodates for this optimization by changing its dictionary lookup method as necessary. First when a dictionary is instantiated, CPython uses a string-only lookup function until a search for non-string is requested, in this case CPython falls back to a more general lookup method and uses this general function onward. There are two main optimizations in the string-only lookup function. First, since the string comparisons do not raise any exceptions by design, some of the error checking logic can been removed. Second, the string-only lookup function can skip the rich set of comparisons ( <=, >=, <, >, ==, !=) which arbitrary data types can provide since string type does not have these special cases.  As it can be seen, string-only keys can dominate the runtime of a dictionary and CPython uses some measures to optimize for this common case. Because of this importance we mostly limit ourselves to the study of new implementations of hashing the string type for the rest of the paper.

The current CPython dictionary implementation uses an iterative XORed polynomial terms for calculating the hash of the strings. In this implementation the hash value for sequential strings differ only in the least significant bits, for instance: hash("6.851")=7824665118634166871 and hash("6.852")=7824665118634166868. This behavior gives good results for sequential strings which is a common case. When the table size is $2^i$ taking the least significant $i$ bits as the index can populate the table without a lot of collisions. This approach gives a better than random results in this case and is also simple and fast to compute.

On the other hand this hashing strategy has the tendency to fill contiguous slots of the
hash table. CPython tries to solve this behavior by utilizing a good collision resolution strategy. The collision resolution strategy currently used is open addressing with a custom non-linear probing scheme. In this scheme CPython uses quadratic probing based on this function: $j = ((5j) + 1) mod 2^i$ where $<0j<2^i-1$. This recurrence alone is not enough for a better behavior since again it behaves the same as linear-probing for some keys since it scans the table entries in a fixed order. To overcome this issue CPython also uses the most significant bits of the hash value with the following code snippet:
\begin{verbatim}
slot = hash;
perturb = hash;
while (<slot is full> && <item in slot does not equal the key>) {
    slot = (5*slot) + 1 + perturb;
    perturb >>= 5;
}
\end{verbatim}
Based on this new method the probe sequence will depend on nearly all the bits of a hash value. We should note that the value of  perturb shift (currently 5) is crucial for a good probing behavior. Since it should be small enough so that  all the significant bits have an  effect on the probe sequence but at the same time it should be large enough so that in really bad cases the high-order hash bits can affect on earlier iterations.
CPython initializes its table with 8 slots. The resizing strategy is based on the load factor invariant. Whenever there are $n$ keys in the dictionary, there must be at least $3n/2$ slots in the table. This load factor is to keep the table sparse and put a bound on the number of collisions that can happen. CPython resizes the size of table whenever the load factor invariant gets violated. It quadruples the size when there are less than 50,000 keys in the table and doubles the size otherwise. 
Overall, the design of CPython dictionary is simple and is optimized for more common cases such as sequential strings. Since it utilizes quadratic probing with perturbation, even if a data type does not provide a good hash function, the table still gets populated with with short chain lengths.  On the other hand there are some behaviors which can be improved. For instance there is a bad cache locality because for probing a slot the table might need to jump to other slots which are not immediate neighbors and not necessarily in cache.

\section{Improvements}
\subsection{Theoretical Results}
Recently, Patrascu et al. [2] showed that simple tabulation hashing provides unexpectedly strong guarantees and is constant competitive with a truly random hash function for hashing with linear probing. They also showed some experimental evaluations which proved that simple tabulation on fixed input length can be fast and robust when used with linear probing and cuckoo hashing.

Linear probing is appealing because of its excellent cache behavior in comparison to the probing scheme currently utilized in Python.  Consider the Intel Core 2 Duo processor with a 64 byte cache line.  Each Python dictionary entry is 12 bytes so when Python’s current probing scheme probes another slot, it is not in the current cache line so it has to pull another cache line in from memory, but with linear probing, that next slot is most likely in the current cache line.

However, Patrascu et al. [2] describe simple tabulation hashing for fixed key widths.  We will need to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).  Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how we implement it.

\subsection{Implementation Details}
We prepopulated random tables in memory with random bits from random.org, which sources randomness form atmospheric noise.

\begin{verbatim}
while (--len >= 0) {
    register long index = (*p++) | ((len&TABLE_MASK)<<8);
    x = x ^ RAND_TABLE_NAME[index];
}
\end{verbatim}

\subsubsection{Optimizations}

Cache

\section{Results}
\subsection{Table Number Tuning}

We attempted to measure the

Eight tables provided the best performance.
\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks to get is to see how the new dictionary performs on when used in each …

This results

Conclusion and future work

Multiple dictionaries, PyPy, etc

Hardware in CPUs
\end{document}

### 2012-05-19 generateAll.py

@@ -2,17 +2,20 @@ import os
 
 versions = {
         'p0': '-DINSTRUMENT_DICT',
-        #'lp': '-DINSTRUMENT_DICT -DLINEAR_PROBING',
+        'lp': '-DINSTRUMENT_DICT -DLINEAR_PROBING',
         #'thm': '-DINSTRUMENT_DICT -DTABULATION_MAIN',
-        'lpthm': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN',
-        'lpths': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT',
-        'lpthspf': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT -DTABULATION_PREFETCH',
+        #'lpthm': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN',
+        #'lpths': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT',
+        #'lpthspf': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT -DTABULATION_PREFETCH',
 #        'lpthmpf': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -fprofile-use',
         #'lpthm1': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=0',
-        #'lpthm2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=1',
-        #'lpthm4': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=3',
-        #'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7',
-        #'lpthm16': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=15',
+#        'lpthm2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=1',
[…]

### 2012-05-19 hash-table-shootout/bench.py

@@ -1,4 +1,4 @@
-import sys, os, subprocess, signal
+import sys, os, subprocess, signal, json
 
 programs = [
 #    'glib_hash_table',
@@ -8,21 +8,24 @@ programs = [
 #    'google_dense_hash_map',
 #    'qt_qhash',
 #    'python_dict.lp',
-    'python_dict.lpthm',
-    'python_dict.lpths',
-    'python_dict.lpthspf',
+#    'python_dict.lpthm',
+#    'python_dict.lpths',
+#    'python_dict.lpthspf',
 #    'python_dict.lpthmpf',
 #    'python_dict.thm',
     'python_dict.p0',
+    'python_dict.lp',
 #    'ruby_hash',
-#    'python_dict.lpthm1',
-#    'python_dict.lpthm2',
-#    'python_dict.lpthm4',
-#    'python_dict.lpthm8',
-#    'python_dict.lpthm16',
+    #'python_dict.lpthm2',
+    #'python_dict.lpthm4',
+    'python_dict.lpthm8',
+    #'python_dict.lpthm16',
+    #'python_dict.lpths2',
+    #'python_dict.lpths4',
+    #'python_dict.lpths8',
 ]
 
-minkeys  =  1*1000*1000
+minkeys  =  2*1000*1000
 #maxkeys  = 40*1000*1000
 maxkeys  = 4*1000*1000
 interval =  2*1000*1000
@@ -57,8 +60,10 @@ for benchtype in benchtypes:
                 # wait for the program to fill up memory and spit out its "ready" message
                 try:
                     runtime = float(proc.stdout.readline().strip())
-                except:
+                    data = json.loads(proc.stdout.readline())
+                except Exception as e:
                     runtime = 0
+                    data = {}
 
                 ps_proc = subprocess.Popen(['ps up %d | tail -n1' % proc.pid], shell=True, stdout=subprocess.PIPE)
                 nbytes = int(ps_proc.stdout.read().split()[4]) * 1024
@@ -68,11 +73,18 @@ for benchtype in benchtypes:
                 proc.wait()
 
                 if nbytes and runtime: # otherwise it crashed
-                    line = ','.join(map(str, [benchtype, nkeys, program, nbytes, "%0.6f" % runtime]))
+                    #line = ','.join(map(str, [benchtype, nkeys, program, nbytes, "%0.6f" % runtime]))
+                    line = data
+                    line['benchtype'] = benchtype
+                    line['nkeys'] = nkeys
[…]

### 2012-05-19 hash-table-shootout/plotoutput.py

@@ -1,30 +1,46 @@
 from collections import defaultdict
 from matplotlib.pyplot import *
+import json
+import matplotlib
+matplotlib.rcParams['lines.linewidth'] = 2
 
-lines = open('output','r').read().split()
-benches = set(map(lambda l: l.split(',')[0], lines))
-dictsizes = sorted(list(set(map(lambda l: int(l.split(',')[1]), lines))))
+datapoints = []
+for line in open('output','r').read().split('\n'):
+    if len(line) > 0:
+        datapoints.append(json.loads(line))
+programs = set(map(lambda l: l['program'], datapoints))
+#benches = set(map(lambda l: l.split(',')[0], lines))
+#dictsizes = sorted(list(set(map(lambda l: int(l.split(',')[1]), lines))))
 #print benches
+#dictsizes = [2,4,6]
+
+lines = set(['python_dict.lpthm8', 'python_dict.lpths4', 'python_dict.lp',
+'python_dict.p0'])
 
 def plot_bench(benchname):
-    filtered_lines = [l for l in lines if l.split(',')[0] == benchname]
+    filtered_points = filter(lambda l: l['benchtype'] == benchname, datapoints)
     results = defaultdict(list)
-    for line in filtered_lines:
-        benchname,size,impl_name,um,time = line.split(',')
-        results[impl_name].append(time)
+    for point in filtered_points:
+#        if point['program'] == 'python_dict.lpthm2':
+#            continue
+        if point['program'] not in lines:
+            continue
+        #results[point['program']].append(point['chain-length'])
+        results[point['program']].append(point['runtime'])
     print results
     ax = figure().add_subplot(111)
 
     for key in results.keys():
-        ax.plot(dictsizes, results[key], label=key)
-    ax.set_xlabel('keys')
+        ax.plot(results[key], label=(key.replace('python_dict.','')))
+    ax.set_xlabel('# keys')
     ax.set_ylabel('time (s)')
-    ax.legend()
+    ax.legend(loc=0)
     return ax
 
-#plot_bench('randomstring')
+plot_bench('sequential')
+title('Sequential Integer Keys')
 #title('Random String Keys')
-plot_bench('sequentialstring')
-title('Sequential String Keys')
+#plot_bench('sequential')
+#title('Sequential String Keys')
 savefig('out.png')
 show()

### 2012-05-19 hash-table-shootout/src/python_dict.c

@@ -19,5 +19,5 @@ typedef PyObject * hash_t;
         PyDict_DelItemString(hash, key); \
     } while(0)
 #define EXIT \
-    PyDict_outputDistribution(hash);
+    printInstrumentDictJsonStats();
 #include "template.c"

### 2012-05-19 hash-table-shootout/src/template.c

@@ -18,7 +18,7 @@ char * new_string_from_integer(int num)
 {
     int ndigits = num == 0 ? 1 : (int)log10(num) + 1;
     char * str = (char *)malloc(ndigits + 1);
-    sprintf(str, "%d", num);
+    sprintf(str, "%d", num, num);
     return str;
 }
 

### 2012-05-19 Objects/stringobject.c
Use hashing with unicode too

@@ -1258,7 +1258,11 @@ _PyString_Eq(PyObject *o1, PyObject *o2)
 }
 
 #ifdef TABULATION_MAIN
+#ifdef TABLE2
+#include "randtable2.c"
+#else
 #include "randtable.c"
+#endif
 #endif
 
 #ifdef TABULATION_SHORT

### 2012-05-19 Objects/unicodeobject.c

@@ -6525,299 +6525,6 @@ unicode_getitem(PyUnicodeObject *self, Py_ssize_t index)
     return (PyObject*) PyUnicode_FromUnicode(&self->str[index], 1);
 }
 
-#ifdef TABULATION
-long
-tabu_hash_unicode(long x)
-{
-    long table0[256] = {
-3823599747470803317, 6049665225936901233, 351709241968265740, -3567193261191146132, -2329246068513330833, -7881953978804000231, -5134155326318512237, 6977199918936832171, \
-3878699803414024627, 6653163622710268196, 3204566177338596744, -1470711584827076289, 117487758559889599, 4438002078960907239, -8649532664808606985, 3961248710215013788, \
-2757123492204357504, 574407754960250196, -6501306440310702141, -4510097668149834244, -2575526590200388415, -7361511802113759183, -2322608073872608665, 194155791575295602, \
--2288486394389341650, -8133757518791266775, -2298015529664186262, 3232145515556952429, 400744677788046308, -1687342448932369124, -4331766055813010854, 8851441406838731884, \
--7325849701687374650, -2425041771848882682, -3037724927721330485, 6626023002520638476, -6711283026153489055, 852388368875775000, -2979248586639516488, 3857807401557377008, \
-8718606972609098740, -6829090877370060631, -3583920417620074790, 254059545051343475, -686536787422719233, 4032380955250148835, -3990778865629872884, 1711256517929123431, \
-156340403005155067, -3665862282139002860, 5873774968511658565, 8091736197025565010, -410816346217722919, 4716048958649107413, 8383048535043026177, 686140283088652783, \
-2731582647533119361, 3212659455936304781, 5221888375944341652, -4702725489219318519, -4983666174764706923, -2747382136344304958, 3314668547721091681, 5582291702751648778, \
--940014190823171282, -7801612559083075502, 8033261866686872982, -6056190462606470467, 4008011538824781130, 974962620650778073, -2255552795693304621, 4195652201914080928, \
-781514889530297078, 6280461301395635219, -5292807593359894832, 5603308539278569319, -4898363640418086628, -4024764785510648649, 6885694195974204527, -3232084626700843404, \
--3063350158683801071, 2546540173480498208, 5038574827895228009, 5261872311948874625, 3852456208346356843, 4434575939837453837, 4233299108678665750, 6003377979757498692, \
[…]

### 2012-05-19 generateAll.py

@@ -11,7 +11,8 @@ versions = {
         #'lpthm1': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=0',
 #        'lpthm2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=1',
 #        'lpthm4': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=3',
-        'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABULATION_PREFETCH',
+        'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7',
+        'lpthm8t2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABLE2',
 #        'lpthm16': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=15',
 #        'lpths2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT -DTABLE_MASK=1',
 #        'lpths4': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT -DTABLE_MASK=3',

### 2012-05-19 hash-table-shootout/bench.py

@@ -19,6 +19,7 @@ programs = [
     #'python_dict.lpthm2',
     #'python_dict.lpthm4',
     'python_dict.lpthm8',
+    'python_dict.lpthm8t2',
     #'python_dict.lpthm16',
     #'python_dict.lpths2',
     #'python_dict.lpths4',

### 2012-05-19 hash-table-shootout/plotoutput.py

@@ -14,7 +14,7 @@ programs = set(map(lambda l: l['program'], datapoints))
 #print benches
 #dictsizes = [2,4,6]
 
-lines = set(['python_dict.lpthm8', 'python_dict.lpths4', 'python_dict.lp',
+lines = set(['python_dict.lpthm8', 'python_dict.lpthm8t2', 'python_dict.lp',
 'python_dict.p0'])
 
 def plot_bench(benchname):
@@ -26,7 +26,8 @@ def plot_bench(benchname):
         if point['program'] not in lines:
             continue
         #results[point['program']].append(point['chain-length'])
-        results[point['program']].append(point['runtime'])
+        #results[point['program']].append(point['runtime'] / int(point['nkeys']))
+        results[point['program']].append(point['scollisioncount'] / int(point['nkeys']))
     print results
     ax = figure().add_subplot(111)
 
@@ -37,7 +38,7 @@ def plot_bench(benchname):
     ax.legend(loc=0)
     return ax
 
-plot_bench('sequential')
+plot_bench('sequentialstring')
 title('Sequential Integer Keys')
 #title('Random String Keys')
 #plot_bench('sequential')

### 2012-05-19 hash-table-shootout/src/template.c

@@ -22,6 +22,14 @@ char * new_string_from_integer(int num)
     return str;
 }
 
+char * new_string_from_long(long num)
+{
+    long ndigits = num == 0 ? 1 : (long)log10(num) + 1;
+    char * str = (char *)malloc(ndigits + 1);
+    sprintf(str, "%ld", num, num);
+    return str;
+}
+
 int main(int argc, char ** argv)
 {
     int num_keys = atoi(argv[1]);
@@ -59,14 +67,14 @@ int main(int argc, char ** argv)
     else if(!strcmp(argv[2], "sequentialstring"))
     {
         for(i = 0; i < num_keys; i++)
-            INSERT_STR_INTO_HASH(new_string_from_integer(i), value);
+            INSERT_STR_INTO_HASH(new_string_from_long(i), value);
     }
 
     else if(!strcmp(argv[2], "randomstring"))
     {
         srandom(1); // for a fair/deterministic comparison
         for(i = 0; i < num_keys; i++)
-            INSERT_STR_INTO_HASH(new_string_from_integer((int)random()), value);
+            INSERT_STR_INTO_HASH(new_string_from_long(random()), value);
     }
 
     else if(!strcmp(argv[2], "deletestring"))

### 2012-05-19 final_paper/final_paper.tex
Use hashing with unicode too

@@
 \begin{abstract}
-This is the paper's abstract \ldots
+Here we implement simple tabulation hashing with linear probing in Python dictionary based on the recent theoretical result of Patrascu et al.
+Although we achieved small performance wins on the order of 4\%, we were unable to achieve significant performance wins.
+The time and memory required to compute the ST hash are too large to justify the speedup garnered by switching to linear probing.
 \end{abstract}
@@
 \section{Introduction}
-In this project we aim to analyze and improve the performance of Python’s dictionary or dict. 
-\paragraph{Outline}
+In this project we aim to analyze and improve the performance of Python’s dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
+1. Clean, elegant near-pseudo code syntax leads to clear code 
+2. Widely distributed and available
+3. Large quantity of external libraries for specific features
 
-\paragraph{Motivation}
+Recently, we the advent of packages like numPy, sciPy and pyPy python has began to also be used (with questionable merit) as a performance computing language due to its ease of use.
+
+However, because Python is an interpreted language it is difficult to operate on a similar performance level to compiled languages due to the sheer overhead of the interpreter.
+
+In order to attempt to improve the performance of the python language as a whole we plan to attempt to improve the performance of python dictionaries.
+
+\subsection{Motivation}
 Python heavily utilizes dictionary data structure for several of its internal language features.
 This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the the user’s program code does not explicitly use a dictionary.
+Take for example, the following code which we ran in the interpreter:
+
+\begin{verbatim}
+        >> i = 2 
+        >> b = i * 2
+            >> b
+            10
+\end{verbatim}
+
+One of the features that we added to the python source was a conditional flag that allowed us view dictionary statistics (collisions, probes, accesses, etc.) for a given invocation of the Python binary.
+With that modification we were able to see that this code used 92 lookups in string dictionaries and 24 lookups in general purpose dictionaries for a total of 116 dictionary hits.
+116 hits in a 3 line piece of code which never uses dictionaries explicitly is quite significant.
+We explain this extensive dictionary usage below.
+
 For instance, a dictionary can be instantiated to pass keyword arguments to a function.
@@
 Some other examples of internal usage of dictionaries in Python are as follows:
-
-Class method lookup
-Instance attribute lookup and global variables lookup
-Python built-ins module
-Membership testing
+1. Class method lookup
+2. Instance attribute lookup and global variables lookup
+3. Python built-ins module
+4. Membership testing
 
@@
 
-\paragraph{Current Python Implementation
-}
+\section{Current Python Implementation}
 
@@
 However as we noted earlier, the dictionaries underlying class instances and modules have only strings as keys.
-Therefore if we can
+Therefore optimizing a dictionary which have string-only keys is crucial to the runtime of any Python program.
+CPython accommodates for this optimization by changing its dictionary lookup method as necessary.
+First when a dictionary is instantiated, CPython uses a string-only lookup function until a search for non-string is requested, in this case CPython falls back to a more general lookup method and uses this general function onward.
+There are two main optimizations in the string-only lookup function.
+First, since the string comparisons do not raise any exceptions by design, some of the error checking logic can been removed.
+Second, the string-only lookup function can skip the rich set of comparisons ( <=, >=, <, >, ==, !=) which arbitrary data types can provide since string type does not have these special cases.
+As it can be seen, string-only keys can dominate the runtime of a dictionary and CPython uses some measures to optimize for this common case.
+Because of this importance we mostly limit ourselves to the study of new implementations of hashing the string type for the rest of the paper.
 
- but contains a specialized string-key-only version for performance in this common case.
-In this project we will focus more on the string data type as the key since most of the mentioned language features use only strings.
-The current Python dictionary implementation uses an iterative XORed polynomial terms for calculating the hash of the strings.
+The current CPython dictionary implementation uses an iterative XORed polynomial terms for calculating the hash of the strings.
+In this implementation the hash value for sequential strings differ only in the least significant bits, for instance: hash("6.851")=7824665118634166871 and hash("6.852")=7824665118634166868.
+This behavior gives good results for sequential strings which is a common case.
+When the table size is $2^i$ taking the least significant $i$ bits as the index can populate the table without a lot of collisions.
+This approach gives a better than random results in this case and is also simple and fast to compute.
+
+On the other hand this hashing strategy has the tendency to fill contiguous slots of the
+hash table.
+CPython tries to solve this behavior by utilizing a good collision resolution strategy.
 The collision resolution strategy currently used is open addressing with a custom non-linear probing scheme.
-We will try to implement a hopefully more performant alternative to this strategy consisting of variants of simple tabulation hashing for variable length strings with linear probing.
+In this scheme CPython uses quadratic probing based on this function: $j = ((5j) + 1) \bmod 2^i$ where $0<j<2^i-1$.
+This recurrence alone is not enough for a better behavior since again it behaves the same as linear-probing for some keys since it scans the table entries in a fixed order.
+To overcome this issue CPython also uses the most significant bits of the hash value with the following code snippet:
+\begin{verbatim}
+slot = hash;
+perturb = hash;
+while (<slot is full> && <item in slot does not equal the key>) {
+       slot = (5*slot) + 1 + perturb;
+       perturb >>= 5;
+}
+\end{verbatim}
+Based on this new method the probe sequence will depend on nearly all the bits of a hash value.
+We should note that the value of  perturb shift (currently 5) is crucial for a good probing behavior.
+Since it should be small enough so that  all the significant bits have an  effect on the probe sequence but at the same time it should be large enough so that in really bad cases the high-order hash bits can affect the earlier iterations.
+CPython initializes its table with 8 slots.
+The resizing strategy is based on a load factor invariant.
+Whenever there are $n$ keys in the dictionary, there must be at least $3n/2$ slots in the table.
+This load factor is to keep the table sparse and put a bound on the number of collisions that can happen.
+CPython resizes the size of table whenever the load factor invariant gets violated.
+It quadruples the size when there are less than 50,000 keys in the table and doubles the size otherwise.
+Overall, the design of CPython dictionary is simple and is optimized for more common cases such as sequential strings.
+Since it utilizes quadratic probing with perturbation, even if a data type does not provide a good hash function, the table still gets populated with with short chain lengths.
+On the other hand there are some behaviors which can be improved.
+For instance there is a bad cache locality because for probing a slot the table might need to jump to the other slots which are not immediate neighbors and not necessarily in cache.
 
-The Theoretical Results and Way to improve implementation
+\section{Improvements}
+\subsection{Theoretical Results}
 Recently, Patrascu et al.
@@
 They also showed some experimental evaluations which proved that simple tabulation on fixed input length can be fast and robust when used with linear probing and cuckoo hashing.
-We plan to implement this scheme and analyze how it performs as a replacement for Python’s dict with the hopes of yielding better performance.
 
-Technical details of implementation
-Main Results
-Benchmarks on real world apps
+Linear probing is appealing because of its excellent cache behavior in comparison to the probing scheme currently utilized in Python.
+Consider the Intel Core 2 Duo processor with a 64 byte cache line.
+Each Python dictionary entry is 12 bytes so when Python’s current probing scheme probes another slot, it is not in the current cache line so it has to pull another cache line in from memory, but with linear probing, that next slot is most likely in the current cache line.
+
+However, Patrascu et al.
+[2] describe simple tabulation hashing for fixed key widths.
+We will need to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).
+Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how we implement it.
+
+\subsection{Implementation Details}
+We prepopulated random tables in memory with random bits from random.org, which sources randomness form atmospheric noise.
+
+       \begin{verbatim}
+    while (--len >= 0) {
+        register long index = (*p++) | ((len&TABLE_MASK)<<8);
+        x = x ^ RAND_TABLE_NAME[index];
+    }
+       \end{verbatim}
+
+\subsubsection{Optimizations}
+
+Cache
+
+\section{Results}
+\subsection{Table Number Tuning}
+
+We attempted to measure the
+
+Eight tables provided the best performance.
+\subsection{Benchmarks}
+In order to analyze the result of this new dictionary we measured three main metrics.
+First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings.
+Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs.
+Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
+Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks is to see how the new dictionary performs on them.
+We used PyBench benchmark suit for this task.
+We saw significant improvements of around \%20 in performance of some of the internals but at the same time some of the internals were running slower compared to current CPython implementation.
+On average there was \%4.5 improvement in runtime for running the whole benchmark suit.
+
+Put some of the number:
+
+PyBench showed the performance difference in each of the individual Python internals however for a more realistic picture of the performance in real world program benchmarking the new implementation was needed.
+Tornado
+
 Conclusion and future work
-
-\section{Previous work}\label{previous work}
-A much longer \LaTeXe{} example was written by Gil~\cite{Gil:02}.
-
-\section{Results}\label{results}
-In this section we describe the results.
-
-\section{Conclusions}\label{conclusions}
-We worked hard, and achieved very little.
-
-\bibliographystyle{abbrv}
-\bibliography{main}
+Hardware specific performance issues dominate at small scales 
+Simple tabulation yields slightly faster and more consistent performance
+Memory latency presents serious but workable issues
+Use different dictionaries for different purposes
+Improve performance bottlenecks illustrated by benchmarks
 

### 2012-05-20 docs/paper.tex

@@
-\documentclass[11pt]{article}
-\title{Improving Python Hashtable Performance with Simple Tabulation and Linear  Probing}
-\author{Russell Cohen, Thomas Georgiou, Pedram Razavi}
+\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
+\author{
+        Russell Cohen \\
+            \and
+        Thomas Georgiou \\
+            \and
+        Pedram Razavi
+}
+\date{\today}
+\documentclass[12pt]{article}
+\usepackage[pdftex]{graphicx}
+
 \begin{document}
 \maketitle
-\section{Abstract} %or is there a \abstract?
+
+\begin{abstract}
 Here we implement simple tabulation hashing with linear probing in Python dictionary based on the recent theoretical result of Patrascu et al.
-Although we achieved small performance wins on the order of 4%, we were unable to achieve significant performance wins.
-The time and memory required to compute the ST hash are too large to justify the speedup garnered by switching to linear probing.
+Although we achieved small performance wins on the order of 4\%, we were unable to achieve significant performance wins.
+The time and memory required to compute the ST hash are too large to justify the speedup garnered by switching to simple tabulation hashing and linear probing.
+\end{abstract}
 
@@
 In this project we aim to analyze and improve the performance of Python’s dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
-1. Clean, elegant near-pseudo code syntax leads to clear code 
-2. Widely distributed and available
-3. Large quantity of external libraries for specific features
+\begin{enumerate}
+ \item Clean, elegant near-pseudo code syntax leads to clear code 
+ \item Widely distributed and available
+ \item Large quantity of external libraries for specific features 
+\end{enumerate}
 
-Recently, we the advent of packages like numPy, sciPy and pyPy python has began to also be used (with questionable merit) as a performance computing language due to its ease of use.
+Recently, with the advent of packages like numPy, sciPy and pyPy Python has began to also be used (with questionable merit) as a performance computing language due to its ease of use.
 
@@
 This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the the user’s program code does not explicitly use a dictionary.
+
 Take for example, the following code which we ran in the interpreter:
@@
 
-One of the features that we added to the python source was a conditional flag that allowed us view dictionary statistics (collisions, probes, accesses, etc.) for a given invocation of the Python binary.
+One of the features that we added to the python source was a conditional flag that allowed us to view dictionary statistics (collisions, probes, accesses, etc.) for a given invocation of the Python binary.
 With that modification we were able to see that this code used 92 lookups in string dictionaries and 24 lookups in general purpose dictionaries for a total of 116 dictionary hits.
@@
 Some other examples of internal usage of dictionaries in Python are as follows:
-1. Class method lookup
-2. Instance attribute lookup and global variables lookup
-3. Python built-ins module
-4. Membership testing
+\begin{itemize}
+\item Class method lookup
+\item Instance attribute lookup and global variables lookup
+\item Python built-ins module
+\item Membership testing
+\end{itemize}
 
@@
 The current CPython dictionary implementation uses an iterative XORed polynomial terms for calculating the hash of the strings.
-In this implementation the hash value for sequential strings differ only in the least significant bits, for instance: hash("6.851")=7824665118634166871 and hash("6.852")=7824665118634166868.
+In this implementation the hash value for sequential strings differ only in the least significant bits, for instance: hash("6.851") = 7824665118634166871 and hash("6.852") = 7824665118634166868.
 This behavior gives good results for sequential strings which is a common case.
@@
 The collision resolution strategy currently used is open addressing with a custom non-linear probing scheme.
-In this scheme CPython uses quadratic probing based on this function: $j = ((5j) + 1) mod 2^i$ where $<0j<2^i-1$.
+In this scheme CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$ where $0<j<2^i-1$.
 This recurrence alone is not enough for a better behavior since again it behaves the same as linear-probing for some keys since it scans the table entries in a fixed order.
@@
 We should note that the value of  perturb shift (currently 5) is crucial for a good probing behavior.
-Since it should be small enough so that  all the significant bits have an  effect on the probe sequence but at the same time it should be large enough so that in really bad cases the high-order hash bits can affect on earlier iterations.
+Since it should be small enough so that  all the significant bits have an  effect on the probe sequence but at the same time it should be large enough so that in really bad cases the high-order hash bits can affect the earlier iterations.
+
 CPython initializes its table with 8 slots.
-The resizing strategy is based on the load factor invariant.
+The resizing strategy is based on a load factor invariant.
 Whenever there are $n$ keys in the dictionary, there must be at least $3n/2$ slots in the table.
@@
 It quadruples the size when there are less than 50,000 keys in the table and doubles the size otherwise.
+
 Overall, the design of CPython dictionary is simple and is optimized for more common cases such as sequential strings.
@@
 On the other hand there are some behaviors which can be improved.
-For instance there is a bad cache locality because for probing a slot the table might need to jump to other slots which are not immediate neighbors and not necessarily in cache.
+For instance there is a bad cache locality because for probing a slot the table might need to jump to the other slots which are not immediate neighbors and not necessarily in cache.
 
 \section{Improvements}
+
 \subsection{Theoretical Results}
@@
 \section{Results}
+
 \subsection{Table Number Tuning}
@@
 Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
-Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks to get is to see how the new dictionary performs on when used in each …
+Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks is to see how the new dictionary performs on them.
+We used PyBench benchmark suit for this task.
+We saw significant improvements of around \%20 in performance of some of the internals such as the special and normal class attribute lookups and string predicates.
+However, at the same time some of the other internals such as CPython exception handler were running around \%12 slower compared to the current implementation.
+On average a \%4.5 improvement in runtime was measures for running the whole benchmark suite.
 
-This results
+[Python bmark Table]
 
-Conclusion and future work
+PyBench showed the performance difference in each of the individual CPython internals however for a more realistic picture of the performance in real world programs more extensive benchmarking of the new implementation was needed.
+Especially since we wanted to know the frequency of the different CPython internals in real-world applications and therefore the corresponding runtime changes.
+The new implementation was tested on several standard real world programs such as Tornado web server, Django web framework, html5lib and nearly 10 other real-world programs.
 
-Multiple dictionaries, PyPy, etc
+We were interested to calculate if the new implementation of dictionary can handle more requests in a Python web server compared to the current CPython implementation.
+We chose Tornado web server for this task since most of the code is written in pure Python.
+An instance of Tornado server was run in a virtual machine with one assigned processor.
+The following code[http://nichol.as/benchmark-of-python-web-servers] was used as a sample Tornado application:
+ \begin{verbatim}
+import tornado.httpserver
+import tornado.ioloop
+import tornado.wsgi
 
-Hardware in CPUs
+def application(environ, start_response):
+    status = '200 OK'
+    output = 'Pong!' 
+    response_headers = [('Content-type', 'text/plain'),
+        ('Content-Length', str(len(output)))]
+    start_response(status, response_headers)
+    return [output]
+def main():
+    container = tornado.wsgi.WSGIContainer(application)
+    http_server = tornado.httpserver.HTTPServer(container)
+    http_server.listen(8000)
+    tornado.ioloop.IOLoop.instance().start()
+if __name__ == "__main__":
+    main()
+ \end{verbatim}
+
+Different request rates (up to 2000 requests per second) were tested on this sample server and their corresponding response rates were calculated using httpef, a HTTP performance measurement tool.
+Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation.
+We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates.
+This finding followed by other benchmarks showed that the new implementation gives more consistent runtime on different iterations of a same code compared to the current CPython implementation.
+This consistency is a desired behavior.
+We suspect that this behavior is a result of the randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.
+
+As mentioned before, another set of the benchmarks that we utilized were based on different standard real-world Python applications.
+We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite.
+The performance of more than ten applications were recoded as shown in the figures.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=160mm]{lr-10.png}
+  \caption{}
+\end{figure}
+
+In a couple of the programs such as pure Python Pickle (an object serialization library) and html5lib (specification parser library) we calculated an improvement of \%3 to \%4.
+In other cases we did not calculate an speedup but at the same time we did not find any real-world case which the new implementation performs significantly slower.
+These results were consistent with the previous results of PyBench that showed about \%4.5 improvement.
+
+\section{Conclusion and future work}
+Throughout the course of our work we came to the following conclusions:
+\begin{enumerate}
+\item Hardware specific performance issues dominate at small scales: When you are attempting to work on a process that runs as fast as a hash table lookup or insertion already does, there is little room for error.
+Improving hash performance with simple tabulation is not an easy task simply because any change you make will probably end up making the hash table run slower.
+That said:
+
+\item Simple tabulation yields slightly faster and more consistent performance: With a careful implementation simple tabulation can yield increased hash performance on linear probing work loads.
+Simple tabulation definitely leads to shorter chain lengths, however, those shorter chain lengths don’t always justify the performance cost.
+
+\item Memory latency presents serious but workable issues.
+Doing a look up in the simple-tabulation table \emph{requires} a random memory access, exactly what we were trying to avoid.
+With careful memory management we can maintain good performance, but a solution requires careful thought about cache sizes and special purpose compiler directives.
+
+\end{enumerate}
+
+In terms of future work, we have 2 avenues of attack.
+First, we could use different dictionaries for different purposes.
+Many of the python dictionaries are small, and their keysets never change.
+No one is adding a new built-in function in the near future.
+Improve performance bottlenecks illustrated by benchmarks
+
 \end{document}

### 2012-05-20 docs/paper.tex
updated paper.tex, removed final_paper.tex
Add in table
added figure benchmark
added improvment graph
added bench graphs

@@
 \usepackage[pdftex]{graphicx}
+\usepackage{float}
 
@@
 We attempted to measure the 
+ \begin{figure}[H]
+   \centering
+   \includegraphics[width=5in]{tables.pdf}
+   \caption{}
+ \end{figure}
 
@@
   \centering
-  \includegraphics[width=160mm]{lr-10.png}
-  \caption{}
+  \includegraphics[width=12cm]{appsbench.pdf}
+  \caption{Prcentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
 \end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=12cm]{slowpickle.png}
+  \caption{Python Pickle module benchmark.
+Minimum Iteration: changed from 0.705s to 0.681s: 1.04 times faster.
+Average iteration: changed from 0.710s to 0.686s: 1.04 times faster.
+Standard deviation of new implementation 2.0383 times smaller}
+\end{figure}
+
+% \begin{figure}[H]
+%   \centering
+%   \includegraphics[width=160mm]{lr-10.png}
+%   \caption{}
+% \end{figure}
+
+% \begin{figure}[H]
+%   \centering
+%   \includegraphics[width=160mm]{lr-10.png}
+%   \caption{}
+% \end{figure}
 

### 2012-05-20 docs/paper.tex
Added pybench table.

--- 2012-05-20 docs/paper.tex
@@
 \date{\today}
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage[pdftex]{graphicx}
+\usepackage{setspace}
+\usepackage{fullpage}
 
@@
 \maketitle
+\doublespace
 
@@
 
-[Python bmark Table]
+\begin{table}
+  \caption{Performance results on selected components of the PyBench suite.  We
+    have ommitted the least significant performace results in terms of percent
+    change.
+  }
+\begin{tabular}{| l | c |}
+  \hline
+  \textbf{Test Name} & \textbf{Performance Difference} \\ \hline
+  Test Name & Performance Difference \\ \hline
+  NormalClassAttribute &  19.20\% \\ \hline
+  SpecialClassAttribute &  18.88\% \\ \hline
+  NormalInstanceAttribute &  17.77\% \\ \hline
+  StringPredicates &  15.25\% \\ \hline
+  UnicodePredicates &  13.82\% \\ \hline
+  SimpleListManipulation &  13.03\% \\ \hline
+  CompareUnicode &  11.27\% \\ \hline
+  SimpleComplexArithmetic &  10.83\% \\ \hline
+  SpecialInstanceAttribute &  10.52\% \\ \hline
+  TryFinally &  10.35\% \\ \hline
+  CreateUnicodeWithConcat &  8.52\% \\ \hline
+  StringSlicing &  6.52\% \\ \hline
+  Small Performance Changes Ommitted & ... \\ \hline
+  UnicodeMappings &  -1.27\% \\ \hline
+  ListSlicing &  -2.40\% \\ \hline
+  ComplexPythonFunctionCalls &  -3.55\% \\ \hline
+  TupleSlicing &  -3.72\% \\ \hline
+  WithRaiseExcept &  -5.89\% \\ \hline
+  TryRaiseExcept &  -11.99\% \\ \hline
+  Totals & 4.54\% \\ \hline
+\end{tabular}
+\end{table}
 
@@
 
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=160mm]{lr-10.png}
-  \caption{}
-\end{figure}
+% \begin{figure}[H]
+%   \centering
+%   \includegraphics[width=160mm]{lr-10.png}
+%   \caption{}
+% \end{figure}
 

### 2012-05-20 docs/paper.tex
Add more things to the paper

--- 2012-05-20 docs/paper.tex
@@
 \date{\today}
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage[pdftex]{graphicx}
@@
 
-Cache
+After implementation and profiling, we found that dictionary table lookups had
+over twice as many level 1 cache misses with tabulation hashing than without.
+Hypothesizing that the tabulation tables were evicting other, more relavant
+data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
+instructions into our hash code to tell the CPU's caching system that it should
+deprioritize storing it in cache.  We used GCC's
+\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
+to do this.  However, this only yielded about a 1\% performance improvement.
 
@@
 We attempted to measure the 
+See Figure ~\ref{fig:tables}
+ \begin{figure}[H]
+   \centering
+   \includegraphics[width=4in]{tables.pdf}
+   \caption{}
+   \label{fig:tables}
+ \end{figure}
 

### 2012-05-20 Objects/stringobject.c

@@ -1304,10 +1304,10 @@ string_hash(PyStringObject *a)
 
 #ifdef TABULATION_MAIN
     while (--len >= 0) {
-        register long index = (*p++) | ((len&TABLE_MASK)<<8);
+        register long index = (*p++) | ((len&TABLE_MASK)<<7);
 #ifdef TABULATION_PREFETCH
-        __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 1);
-        __builtin_prefetch(p, 0, 1);
+        __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 2);
+//        __builtin_prefetch(p, 0, 0);
 #endif
         x = x ^ RAND_TABLE_NAME[index];  //assume we are on a 64bit machine
         //printf("%lx\n",x);

### 2012-05-20 generateAll.py

@@ -12,6 +12,7 @@ versions = {
 #        'lpthm2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=1',
 #        'lpthm4': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=3',
         'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7',
+        'lpthm8pf': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABULATION_PREFETCH',
         'lpthm8t2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABLE2',
 #        'lpthm16': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=15',
 #        'lpths2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT -DTABLE_MASK=1',

### 2012-05-20 docs/paper.tex

@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=12cm]{appsbench.pdf}
+  \caption{Prcentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=12cm]{slowpickle.png}
+  \caption{Python Pickle module benchmark.
+Minimum Iteration: changed from 0.705s to 0.681s: 1.04 times faster.
+Average iteration: changed from 0.710s to 0.686s: 1.04 times faster.
+Standard deviation of new implementation 2.0383 times smaller}
+\end{figure}
+
+% \begin{figure}[H]
+%   \centering
+%   \includegraphics[width=160mm]{lr-10.png}
+%   \caption{}
+% \end{figure}
+
 % \begin{figure}[H]

### 2012-05-20 docs/paper.tex

--- 2012-05-20 docs/paper.tex
@@
 \usepackage{fullpage}
+\usepackage{float}
 
@@
 
-One of the features that we added to the python source was a conditional flag that allowed us to view dictionary statistics (collisions, probes, accesses, etc.) for a given invocation of the Python binary.
-With that modification we were able to see that this code used 92 lookups in string dictionaries and 24 lookups in general purpose dictionaries for a total of 116 dictionary hits.
-116 hits in a 3 line piece of code which never uses dictionaries explicitly is quite significant.
-We explain this extensive dictionary usage below.
+One of the features that we added to the python source was a conditional flag
+that allowed us to view dictionary statistics (collisions, probes, accesses,
+etc.) for a given invocation of the Python binary.  With that modification we
+were able to see that this code used 92 lookups in string dictionaries and 24
+lookups in general purpose dictionaries for a total of 116 dictionary hits. 116
+hits in a 3 line piece of code which never uses dictionaries explicitly is quite
+significant.  We explain this extensive dictionary usage below.
 
-For instance, a dictionary can be instantiated to pass keyword arguments to a function.
-This means a dictionary could potentially be created and destroyed on every function call.
-Some other examples of internal usage of dictionaries in Python are as follows:
+For instance, a dictionary can be instantiated to pass keyword arguments to a
+function. This means a dictionary could potentially be created and destroyed on
+every function call.  Some other examples of internal usage of dictionaries in
+Python are as follows:
+
 \begin{itemize}
@@
 
-Later in this paper, we compare the performance of our new dictionary implementation for some of these special cases.
-The aforementioned applications of the dictionaries further highlights the need for the fast instantiation and deletion of dictionaries so that less memory is utilized once running a program.
-Aside from the issue of memory usage, to further enhance the runtime of a program we need fast key/value storage and retrievals which is the main focus of this paper.
-Therefore it becomes clear that a better dictionary structure will have significant effects on total Python performance.
+Later in this paper, we compare the performance of our new dictionary
+implementation for some of these special cases. The aforementioned applications
+of the dictionaries further highlights the need for the fast instantiation and
+deletion of dictionaries so that less memory is utilized once running a program.
+Aside from the issue of memory usage, to further enhance the runtime of a
+program we need fast key/value storage and retrievals which is the main focus of
+this paper. Therefore it becomes clear that a better dictionary structure will
+have significant effects on total Python performance.
 
@@
 
-In this section we give a brief overview of the key implementation details of dictionaries in CPython 2.7.3.
-We should note that Python’s dictionary supports several different data types as keys.
-However as we noted earlier, the dictionaries underlying class instances and modules have only strings as keys.
-Therefore optimizing a dictionary which have string-only keys is crucial to the runtime of any Python program.
-CPython accommodates for this optimization by changing its dictionary lookup method as necessary.
-First when a dictionary is instantiated, CPython uses a string-only lookup function until a search for non-string is requested, in this case CPython falls back to a more general lookup method and uses this general function onward.
-There are two main optimizations in the string-only lookup function.
-First, since the string comparisons do not raise any exceptions by design, some of the error checking logic can been removed.
-Second, the string-only lookup function can skip the rich set of comparisons ( <=, >=, <, >, ==, !=) which arbitrary data types can provide since string type does not have these special cases.
-As it can be seen, string-only keys can dominate the runtime of a dictionary and CPython uses some measures to optimize for this common case.
-Because of this importance we mostly limit ourselves to the study of new implementations of hashing the string type for the rest of the paper.
+In this section we give a brief overview of the key implementation details of
+dictionaries in CPython 2.7.3.  We should note that Python’s dictionary supports
+several different data types as keys. However as we noted earlier, the
+dictionaries underlying class instances and modules have only strings as keys.
+Therefore optimizing a dictionary which have string-only keys is crucial to the
+runtime of any Python program. CPython accommodates for this optimization by
+changing its dictionary lookup method as necessary. First when a dictionary is
+instantiated, CPython uses a string-only lookup function until a search for
+non-string is requested, in this case CPython falls back to a more general
+lookup method and uses this general function onward. There are two main
+optimizations in the string-only lookup function.  First, since the string
+comparisons do not raise any exceptions by design, some of the error checking
+logic can been removed. Second, the string-only lookup function can skip the
+rich set of comparisons ( $<=$, $>=$, $<$, $>$, $==$, $!=$) which arbitrary data
+types can provide since string type does not have these special cases.  As it
+can be seen, string-only keys can dominate the runtime of a dictionary and
+CPython uses some measures to optimize for this common case. Because of this
+importance we mostly limit ourselves to the study of new implementations of
+hashing the string type for the rest of the paper.
 
-The current CPython dictionary implementation uses an iterative XORed polynomial terms for calculating the hash of the strings.
-In this implementation the hash value for sequential strings differ only in the least significant bits, for instance: hash("6.851") = 7824665118634166871 and hash("6.852") = 7824665118634166868.
-This behavior gives good results for sequential strings which is a common case.
-When the table size is $2^i$ taking the least significant $i$ bits as the index can populate the table without a lot of collisions.
-This approach gives a better than random results in this case and is also simple and fast to compute.
+The current CPython dictionary implementation uses an iterative XORed polynomial
+terms for calculating the hash of the strings. In this implementation the hash
+value for sequential strings differ only in the least significant bits, for
+instance: hash("6.851") = 7824665118634166871 and hash("6.852") =
+7824665118634166868. This behavior gives good results for sequential strings
+which is a common case. When the table size is $2^i$ taking the least
+significant $i$ bits as the index can populate the table without a lot of
+collisions. This approach gives a better than random results in this case and is
+also simple and fast to compute.
 
-On the other hand this hashing strategy has the tendency to fill contiguous slots of the
-hash table.
-CPython tries to solve this behavior by utilizing a good collision resolution strategy.
-The collision resolution strategy currently used is open addressing with a custom non-linear probing scheme.
-In this scheme CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$ where $0<j<2^i-1$.
-This recurrence alone is not enough for a better behavior since again it behaves the same as linear-probing for some keys since it scans the table entries in a fixed order.
-To overcome this issue CPython also uses the most significant bits of the hash value with the following code snippet:
+On the other hand this hashing strategy has the tendency to fill contiguous
+slots of the hash table. CPython tries to solve this behavior by utilizing a
+good collision resolution strategy. The collision resolution strategy currently
+used is open addressing with a custom non-linear probing scheme. In this scheme
+CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$
+where $0<j<2^i-1$. This recurrence alone is not enough for a better behavior
+since again it behaves the same as linear-probing for some keys since it scans
+the table entries in a fixed order. To overcome this issue CPython also uses the
+most significant bits of the hash value with the following code snippet:
+
 \begin{verbatim}
@@
 \end{verbatim}
-Based on this new method the probe sequence will depend on nearly all the bits of a hash value.
-We should note that the value of  perturb shift (currently 5) is crucial for a good probing behavior.
-Since it should be small enough so that  all the significant bits have an  effect on the probe sequence but at the same time it should be large enough so that in really bad cases the high-order hash bits can affect the earlier iterations.
 
-CPython initializes its table with 8 slots.
-The resizing strategy is based on a load factor invariant.
-Whenever there are $n$ keys in the dictionary, there must be at least $3n/2$ slots in the table.
-This load factor is to keep the table sparse and put a bound on the number of collisions that can happen.
-CPython resizes the size of table whenever the load factor invariant gets violated.
-It quadruples the size when there are less than 50,000 keys in the table and doubles the size otherwise.
+Based on this new method the probe sequence will depend on nearly all the bits
+of a hash value. We should note that the value of  perturb shift (currently 5)
+is crucial for a good probing behavior. Since it should be small enough so that
+all the significant bits have an  effect on the probe sequence but at the same
+time it should be large enough so that in really bad cases the high-order hash
+bits can affect the earlier iterations.
 
-Overall, the design of CPython dictionary is simple and is optimized for more common cases such as sequential strings.
-Since it utilizes quadratic probing with perturbation, even if a data type does not provide a good hash function, the table still gets populated with with short chain lengths.
-On the other hand there are some behaviors which can be improved.
-For instance there is a bad cache locality because for probing a slot the table might need to jump to the other slots which are not immediate neighbors and not necessarily in cache.
+CPython initializes its table with 8 slots. The resizing strategy is based on a
+load factor invariant. Whenever there are $n$ keys in the dictionary, there must
+be at least $3n/2$ slots in the table. This load factor is to keep the table
+sparse and put a bound on the number of collisions that can happen. CPython
+resizes the size of table whenever the load factor invariant gets violated. It
+quadruples the size when there are less than 50,000 keys in the table and
+doubles the size otherwise.
+
+Overall, the design of CPython dictionary is simple and is optimized for more
+common cases such as sequential strings. Since it utilizes quadratic probing
+with perturbation, even if a data type does not provide a good hash function,
+the table still gets populated with with short chain lengths.  On the other hand
+there are some behaviors which can be improved. For instance there is a bad
+cache locality because for probing a slot the table might need to jump to the
+other slots which are not immediate neighbors and not necessarily in cache.
 

### 2012-05-20 docs/paper.tex

@@
 We attempted to measure the 
+See Figure ~\ref{fig:tables}
  \begin{figure}[H]
    \centering
-   \includegraphics[width=5in]{tables.pdf}
+   \includegraphics[width=4in]{tables.pdf}
    \caption{}
+   \label{fig:tables}
  \end{figure}

### 2012-05-20 docs/paper.tex
update the theoretical results

--- 2012-05-20 docs/paper.tex
@@
 \date{\today}
-\documentclass[11pt]{article}
+\documentclass[12pt]{article}
 \usepackage[pdftex]{graphicx}
@@
 [2] showed that simple tabulation hashing provides unexpectedly strong guarantees and is constant competitive with a truly random hash function for hashing with linear probing.
+In their paper, they showed that for simple tabulation hashing with linear probing if the table size is $m = (1 + \epsilon)n$, then the expected time per operation is $O(1/\epsilon^2)$.
+This new guarantee asymptotically matches the
+bound for a truly random function.
 They also showed some experimental evaluations which proved that simple tabulation on fixed input length can be fast and robust when used with linear probing and cuckoo hashing.
@@
 [2] describe simple tabulation hashing for fixed key widths.
-We will need to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).
-Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how we implement it.
+For implementing this new hashing scheme, first we needed to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).
+Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how to implement this.
+In the following section we explain the implementation details of our approach to these problems.
 
@@
 
-After implementation and profiling, we found that dictionary table lookups had
-over twice as many level 1 cache misses with tabulation hashing than without.
-Hypothesizing that the tabulation tables were evicting other, more relavant
-data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
-instructions into our hash code to tell the CPU's caching system that it should
-deprioritize storing it in cache.  We used GCC's
-\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
-to do this.  However, this only yielded about a 1\% performance improvement.
+Cache
 
@@
 We attempted to measure the 
-See Figure ~\ref{fig:tables}
  \begin{figure}[H]
    \centering
-   \includegraphics[width=4in]{tables.pdf}
+   \includegraphics[width=5in]{tables.pdf}
    \caption{}
-   \label{fig:tables}
  \end{figure}

### 2012-05-20 docs/paper.tex
moved tornado code to appendix

@@
 An instance of Tornado server was run in a virtual machine with one assigned processor.
-The following code[http://nichol.as/benchmark-of-python-web-servers] was used as a sample Tornado application:
- \begin{verbatim}
-import tornado.httpserver
-import tornado.ioloop
-import tornado.wsgi
-
-def application(environ, start_response):
-    status = '200 OK'
-    output = 'Pong!' 
-    response_headers = [('Content-type', 'text/plain'),
-        ('Content-Length', str(len(output)))]
-    start_response(status, response_headers)
-    return [output]
-def main():
-    container = tornado.wsgi.WSGIContainer(application)
-    http_server = tornado.httpserver.HTTPServer(container)
-    http_server.listen(8000)
-    tornado.ioloop.IOLoop.instance().start()
-if __name__ == "__main__":
-    main()
- \end{verbatim}
+The code of the application used is provided in the appendix.
 
@@
 
+\appendix
+\section{Tornado Application}
+Sample Tornado 2.2.1 application used for benchmark:
+ \begin{verbatim}
+import tornado.httpserver
+import tornado.ioloop
+import tornado.wsgi
+
+def application(environ, start_response):
+    status = '200 OK'
+    output = 'Hello World!' 
+    response_headers = [('Content-type', 'text/plain'),
+        ('Content-Length', str(len(output)))]
+    start_response(status, response_headers)
+    return [output]
+def main():
+    container = tornado.wsgi.WSGIContainer(application)
+    http_server = tornado.httpserver.HTTPServer(container)
+    http_server.listen(8000)
+    tornado.ioloop.IOLoop.instance().start()
+if __name__ == "__main__":
+    main()
+ \end{verbatim}
+\appendix
+
 \end{document}

### 2012-05-20 docs/paper.tex
Writing + more benchmarks

--- 2012-05-20 docs/paper.tex
@@
 
-We attempted to measure the 
-See Figure ~\ref{fig:tables}
+For simple tabulation hashing with variable length strings, we decided to use a
+fixed number of tables $N$ and round robin between them for every byte.  We
+found the optimal value of $N$ to be 8 experimentally.  We limited our choices
+to powers of 2 so that costly modulo operations could be avoided by using bit
+masks Figure ~\ref{fig:tables} shows the performance of different numbers of
+tabulation tables.  Eight turned out to be the sweetspot
  \begin{figure}[H]
@@
    \includegraphics[width=4in]{tables.pdf}
-   \caption{}
+   \caption{6000000 keys}
    \label{fig:tables}

### 2012-05-20 Objects/stringobject.c

@@ -1304,7 +1304,7 @@ string_hash(PyStringObject *a)
 
 #ifdef TABULATION_MAIN
     while (--len >= 0) {
-        register long index = (*p++) | ((len&TABLE_MASK)<<7);
+        register long index = (*p++) | ((len&TABLE_MASK)<<8);
 #ifdef TABULATION_PREFETCH
         __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 2);
 //        __builtin_prefetch(p, 0, 0);

### 2012-05-20 generateAll.py

@@ -3,6 +3,7 @@ import os
 versions = {
         'p0': '-DINSTRUMENT_DICT',
         'lp': '-DINSTRUMENT_DICT -DLINEAR_PROBING',
+        'thm8': '-DINSTRUMENT_DICT -DTABULATION_MAIN',
         #'thm': '-DINSTRUMENT_DICT -DTABULATION_MAIN',
         #'lpthm': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN',
         #'lpths': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_SHORT',

### 2012-05-20 hash-table-shootout/bench.py

@@ -13,13 +13,14 @@ programs = [
 #    'python_dict.lpthspf',
 #    'python_dict.lpthmpf',
 #    'python_dict.thm',
-#    'python_dict.p0',
-#    'python_dict.lp',
+    'python_dict.p0',
+    'python_dict.lp',
+    'python_dict.thm8',
 #    'ruby_hash',
     #'python_dict.lpthm2',
     #'python_dict.lpthm4',
     'python_dict.lpthm8',
-    'python_dict.lpthm8pf',
+#    'python_dict.lpthm8pf',
 #    'python_dict.lpthm8t2',
     #'python_dict.lpthm16',
     #'python_dict.lpths2',
@@ -27,9 +28,9 @@ programs = [
     #'python_dict.lpths8',
 ]
 
-minkeys  =  2*1000*1000
+minkeys  =  1*1000*1000
 #maxkeys  = 40*1000*1000
-maxkeys  = 4*1000*1000
+maxkeys  = 2*1000*1000
 interval =  2*1000*1000
 best_out_of = 3
 

### 2012-05-20 hash-table-shootout/src/template.c

@@ -25,8 +25,8 @@ char * new_string_from_integer(int num)
 char * new_string_from_long(long num)
 {
     long ndigits = num == 0 ? 1 : (long)log10(num) + 1;
-    char * str = (char *)malloc(ndigits + 1);
-    sprintf(str, "%ld", num, num);
+    char * str = (char *)malloc(2*ndigits + 1);
+    sprintf(str, "%ld %ld", num, num);
     return str;
 }
 

### 2012-05-20 Objects/stringobject.c
Chagnes...

@@ -1306,7 +1306,7 @@ string_hash(PyStringObject *a)
     while (--len >= 0) {
         register long index = (*p++) | ((len&TABLE_MASK)<<8);
 #ifdef TABULATION_PREFETCH
-        __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 2);
+        __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 1);
 //        __builtin_prefetch(p, 0, 0);
 #endif
         x = x ^ RAND_TABLE_NAME[index];  //assume we are on a 64bit machine

### 2012-05-20 generateAll.py

@@ -12,7 +12,7 @@ versions = {
         #'lpthm1': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=0',
 #        'lpthm2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=1',
 #        'lpthm4': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=3',
-        'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7',
+        'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABLE2',
         'lpthm8pf': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABULATION_PREFETCH',
         'lpthm8t2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABLE2',
 #        'lpthm16': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=15',

### 2012-05-20 Objects/stringobject.c
Newer hash table shootout benchmark + fix  zeroing out due to xor

@@ -1309,7 +1309,7 @@ string_hash(PyStringObject *a)
         __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 1);
 //        __builtin_prefetch(p, 0, 0);
 #endif
-        x = x ^ RAND_TABLE_NAME[index];  //assume we are on a 64bit machine
+        x = x ^ (RAND_TABLE_NAME[index] >> len);  //assume we are on a 64bit machine
         //printf("%lx\n",x);
     }
 #else

### 2012-05-20 generateAll.py

@@ -12,7 +12,7 @@ versions = {
         #'lpthm1': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=0',
 #        'lpthm2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=1',
 #        'lpthm4': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=3',
-        'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABLE2',
+        'lpthm8': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7',
         'lpthm8pf': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABULATION_PREFETCH',
         'lpthm8t2': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=7 -DTABLE2',
 #        'lpthm16': '-DINSTRUMENT_DICT -DLINEAR_PROBING -DTABULATION_MAIN -DTABLE_MASK=15',

### 2012-05-20 hash-table-shootout/bench.py

@@ -20,7 +20,7 @@ programs = [
     #'python_dict.lpthm2',
     #'python_dict.lpthm4',
     'python_dict.lpthm8',
-#    'python_dict.lpthm8pf',
+    'python_dict.lpthm8pf',
 #    'python_dict.lpthm8t2',
     #'python_dict.lpthm16',
     #'python_dict.lpths2',
@@ -28,9 +28,9 @@ programs = [
     #'python_dict.lpths8',
 ]
 
-minkeys  =  1*1000*1000
+minkeys  =  2*1000*1000
 #maxkeys  = 40*1000*1000
-maxkeys  = 2*1000*1000
+maxkeys  = 6*1000*1000
 interval =  2*1000*1000
 best_out_of = 3
 

### 2012-05-20 docs/paper.tex
Add graphs

@@
 \subsection{Implementation Details}
-We prepopulated random tables in memory with random bits from random.org, which sources randomness form atmospheric noise.
+We prepopulated random tables in memory with random bits from random.org, which
+sources randomness form atmospheric noise in order to remove the reliance on
+pseudo-random number generators.
+
+      Following is an excerpt from our hashing code.  We use bit tricks in
+      order to prevent costly modulo instructions.  For each byte in the
+      string, we construct an index by adding the byte value and a table offset
+      together.  Then we lookup the value in a large table of random 64 bit
+      longs and then XOR it with the previous result in the chain of XORs.
 

### 2012-05-20 Objects/dictobject.c

@@ -845,7 +845,7 @@ PyDict_SetItem(register PyObject *op, PyObject *key, PyObject *value)
      * Very large dictionaries (over 50K items) use doubling instead.
      * This may help applications with severe memory constraints.
      */
-    if (!(mp->ma_used > n_used && mp->ma_fill*4 >= (mp->ma_mask+1)*2))
+    if (!(mp->ma_used > n_used && mp->ma_fill*3 >= (mp->ma_mask+1)*2))
         return 0;
     return dictresize(mp, (mp->ma_used > 50000 ? 2 : 4) * mp->ma_used);
 }

### 2012-05-20 hash-table-shootout/plotoutput.py

@@ -14,7 +14,7 @@ programs = set(map(lambda l: l['program'], datapoints))
 #print benches
 #dictsizes = [2,4,6]
 
-lines = set(['python_dict.lpthm8', 'python_dict.lpthm8t2', 'python_dict.lp',
+lines = set(['python_dict.lpthm8','python_dict.lpthm8pf', 'python_dict.thm8', 'python_dict.lp',
 'python_dict.p0'])
 
 def plot_bench(benchname):
@@ -27,7 +27,8 @@ def plot_bench(benchname):
             continue
         #results[point['program']].append(point['chain-length'])
         #results[point['program']].append(point['runtime'] / int(point['nkeys']))
-        results[point['program']].append(point['scollisioncount'] / int(point['nkeys']))
+        #results[point['program']].append(point['scollisioncount'] / int(point['nkeys']))
+        results[point['program']].append(point['runtime'])
     print results
     ax = figure().add_subplot(111)
 

### 2012-05-20 docs/paper.tex
Fix spelling errors

@@
 \section{Introduction}
-In this project we aim to analyze and improve the performance of Python’s dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
+In this project we aim to analyze and improve the performance of Python's dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
 \begin{enumerate}
@@
 Python heavily utilizes dictionary data structure for several of its internal language features.
-This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the the user’s program code does not explicitly use a dictionary.
+This extensive internal use of dictionaries means that a running Python program has many dictionaries active simultaneously, even in cases where the user's program code does not explicitly use a dictionary.
 
@@
 In this section we give a brief overview of the key implementation details of
-dictionaries in CPython 2.7.3.  We should note that Python’s dictionary supports
+dictionaries in CPython 2.7.3.  We should note that Python's dictionary supports
 several different data types as keys. However as we noted earlier, the
@@
 over twice as many level 1 cache misses with tabulation hashing than without.
-Hypothesizing that the tabulation tables were evicting other, more relavant
+Hypothesizing that the tabulation tables were evicting other, more relevant
 data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
@@
 masks Figure ~\ref{fig:tables} shows the performance of different numbers of
-tabulation tables.  Eight turned out to be the sweetspot
+tabulation tables.  Eight turned out to be optimal.
  \begin{figure}[H]
@@
 Eight tables provided the best performance.
+
+ \begin{figure}[H]
+   \centering
+   \includegraphics[width=4in]{randstring.pdf}
+   \caption{Random string performance}
+   \label{fig:randstring}
+ \end{figure}
+ \begin{figure}[H]
+   \centering
+   \includegraphics[width=4in]{seqstring.pdf}
+   \caption{Sequential string performance}
+   \label{fig:seqstring}
+ \end{figure}
+
 \subsection{Benchmarks}
@@
   \caption{Performance results on selected components of the PyBench suite.  We
-    have ommitted the least significant performace results in terms of percent
+    have omitted the least significant performance results in terms of percent
     change.
@@
   \textbf{Test Name} & \textbf{Performance Difference} \\ \hline
-  Test Name & Performance Difference \\ \hline
   NormalClassAttribute &  19.20\% \\ \hline
@@
   StringSlicing &  6.52\% \\ \hline
-  Small Performance Changes Ommitted & ... \\ \hline
+  Small Performance Changes Omitted & ... \\ \hline
   UnicodeMappings &  -1.27\% \\ \hline
@@
 
-Different request rates (up to 2000 requests per second) were tested on this sample server and their corresponding response rates were calculated using httpef, a HTTP performance measurement tool.
+Different request rates (up to 2000 requests per second) were tested on this
+sample server and their corresponding response rates were calculated using
+httperf, a HTTP performance measurement tool.
 Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation.
@@
   \includegraphics[width=12cm]{appsbench.pdf}
-  \caption{Prcentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
+  \caption{Percentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
 \end{figure}

### 2012-05-20 docs/paper.tex
Add in figure to the paper
Font changes
Add in hardware randomness blurb

@@
 \maketitle
-\doublespace
+%\doublespace
 
@@
  \end{figure}
+ \begin{figure}[H]
+   \centering
+   \includegraphics[width=4in]{chainlength.pdf}
+   \caption{Simple tabulation hashing decreases the average chain length}
+   \label{fig:seqstring}
+ \end{figure}
 
@@
 
+Given the strong performance of simple tabulation hashing and the only downside
+being the cost of memory lookups, processor manufacturers should look into
+adding random tables into hardware as a ROM.  We found that for what we found
+optimal for Python, we only needed 16KB of randomness.  This would be a trivial
+cost in ROM given that processors have MBs of higher level caches nowadays.
+
 \appendix

### 2012-05-20 Objects/stringobject.c
Bugfixes

@@ -1309,7 +1309,7 @@ string_hash(PyStringObject *a)
         __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 1);
 //        __builtin_prefetch(p, 0, 0);
 #endif
-        x = x ^ (RAND_TABLE_NAME[index] >> len);  //assume we are on a 64bit machine
+        x = x ^ (RAND_TABLE_NAME[index] >> (len&63));  //assume we are on a 64bit machine
         //printf("%lx\n",x);
     }
 #else

### 2012-05-20 Objects/unicodeobject.c

@@ -6577,7 +6577,7 @@ unicode_hash(PyUnicodeObject *self)
         __builtin_prefetch(&RAND_TABLE_NAME[index], 0, 1);
         __builtin_prefetch(p, 0, 1);
 #endif
-        x = x ^ RAND_TABLE_NAME[index];  //assume we are on a 64bit machine
+        x = x ^ (RAND_TABLE_NAME[index] >> (len&63));  //assume we are on a 64bit machine
         //printf("%lx\n",x);
     }
 #else

### 2012-05-20 docs/paper.tex
More implementation details
More writing

@@
 
+We also experimented with indexing using every pair of bytes in the string and
+using $2^16$ tabulation table entries.
+
+To implement linear probing, we modified Python's dictionary code to start at
+the table entry defined by the hash value and then search sequentially until it
+finds the item it is looking for or an empty cell.
 \subsubsection{Optimizations}
@@
 \begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
-to do this.  However, this only yielded about a 1\% performance improvement.
+to do this.  However, this only yielded about a 1\% performance improvement,
+negligible compared to the performance improvement of different hash schemes.
 
@@
 masks Figure ~\ref{fig:tables} shows the performance of different numbers of
-tabulation tables.  Eight turned out to be optimal.
+tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
+every byte separately resulted in the best performance.    We believe this is due
+to the low entropy for fewer tables and too much memory bandwidth for a higher
+number of tables.  The optimal result may also be an artifact of the fact that
+we were using on the order of magnitude of thousands to millions of dictionary
+keys.  Our optimal result resulted 256 64
+bit table entries in each of 8 tables using 16KB of memory.
  \begin{figure}[H]
    \centering
-   \includegraphics[width=4in]{tables.pdf}
-   \caption{6000000 keys}
+   \includegraphics[width=4.1in]{tables.pdf}
+   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
+   dictionary with 6000000 keys}
    \label{fig:tables}
@@
 
-Eight tables provided the best performance.
-
  \begin{figure}[H]
    \centering
-   \includegraphics[width=4in]{randstring.pdf}
+   \includegraphics[width=4.1in]{randstring.pdf}
    \caption{Random string performance}
@@
    \centering
-   \includegraphics[width=4in]{seqstring.pdf}
+   \includegraphics[width=4.1in]{seqstring.pdf}
    \caption{Sequential string performance}
@@
    \centering
-   \includegraphics[width=4in]{chainlength.pdf}
+   \includegraphics[width=4.1in]{chainlength.pdf}
    \caption{Simple tabulation hashing decreases the average chain length}

### 2012-05-20 docs/paper.tex
Explain hash table shootout

@@
    \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
-   dictionary with 6000000 keys}
+   dictionary with 6 million keys}
    \label{fig:tables}
  \end{figure}
+
+\subsection{Synthetic Benchmarks}
+
+Our primary benchmark was a synthetic one that called the CPython dictionary
+code directly from C so that all the interpreter overhead was bypassed.  We
+measured performance for both sequential strings, generated by converting
+sequential integers into their string representation, and for random strings,
+generated by the string representation of random integers.  For each of the
+benchmarks, we inserted a given number of keys into the dictionary and measured
+the runtime.
+
+Figure ~\ref{fig:randstring} shows the performance for random strings.  Both
+linear probing and simple tabulation result in improved performance over the
+current CPython dictionary implementation but combining both techniques results
+in over a 20\% improvement.  Figure ~\ref{fig:seqstring} shows very similar
+results for sequential strings.
 
@@
 \end{figure}
+
+Figure ~\ref{fig:chainlength} shows the average chain length, or dictionary
+cell probes per lookup for a run of the benchmark on sequential strings.
+Simple tabulation hashing with CPython's current collision resolution strategy
+results in the fewest number of probes.  However, despite having the fewest
+number of probes, its real relative performance is much worse as seen in Figure
+~\ref{fig:seqstring}.  Linear probing performs faster than its average
+chain length suggests, showing the impact of looking at entries in the same
+cache line first rather than jumping around.
+
 \begin{figure}[H]
@@
   \caption{Simple tabulation hashing decreases the average chain length}
-   \label{fig:seqstring}
+  \label{fig:chainlength}
 \end{figure}
@@
 Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
-Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks is to see how the new dictionary performs on them.
-We used PyBench benchmark suit for this task.
-We saw significant improvements of around \%20 in performance of some of the internals such as the special and normal class attribute lookups and string predicates.
-However, at the same time some of the other internals such as CPython exception handler were running around \%12 slower compared to the current implementation.
-On average a \%4.5 improvement in runtime was measures for running the whole benchmark suite.
+Because of the heavy utilization of dictionaries in Python internals one of the
+valuable benchmarks is to see how the new dictionary performs on them. We used
+PyBench benchmark suit for this task. We saw significant improvements of around
+20\% in performance of some of the internals such as the special and normal
+class attribute lookups and string predicates. However, at the same time some
+of the other internals such as CPython exception handler were running around
+12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measures for running the whole benchmark suite.
 
@@
 
-In a couple of the programs such as pure Python Pickle (an object serialization library) and html5lib (specification parser library) we calculated an improvement of \%3 to \%4.
-In other cases we did not calculate an speedup but at the same time we did not find any real-world case which the new implementation performs significantly slower.
-These results were consistent with the previous results of PyBench that showed about \%4.5 improvement.
+In a couple of the programs such as pure Python Pickle (an object serialization
+library) and html5lib (specification parser library) we calculated an
+improvement of 3\% to 4\%. In other cases we did not calculate an speedup but
+at the same time we did not find any real-world case which the new
+implementation performs significantly slower. These results were consistent
+with the previous results of PyBench that showed about 4.5\% improvement.
 

### 2012-05-20 docs/paper.tex
Capitalization of python

@@
 \begin{abstract}
-Here we implement simple tabulation hashing with linear probing in Python dictionary based on the recent theoretical result of Patrascu et al.
-Although we achieved small performance wins on the order of 4\%, we were unable to achieve significant performance wins.
-The time and memory required to compute the ST hash are too large to justify the speedup garnered by switching to simple tabulation hashing and linear probing.
+Here we implement simple tabulation hashing with linear probing in Python
+dictionary based on the recent theoretical result of Patrascu et al. Although
+we achieved significant performance wins on synthetic benchmarks, on real
+benchmarks our wins were only on the order of 4\%.  When viewed as a whole, the
+extra memory bandwidth
+and latency required to compute the simple tabulation hash are too large to
+justify the dictionary speedup garnered by switching to simple tabulation hashing and linear probing.
 \end{abstract}
@@
 \section{Introduction}
-In this project we aim to analyze and improve the performance of Python's dictionary or dict.  Python is an extremely popular general purpose computing language for several reasons:
+In this project we aim to analyze and improve the performance of Python's
+dictionary or dict by changing its underlying implementation.  Python is an extremely popular general purpose computing language for several reasons:
 \begin{enumerate}
@@
 
-Our primary benchmark was a synthetic one that called the CPython dictionary
+Our primary benchmark, hash-table-shootout, is a synthetic one that called the CPython dictionary
 code directly from C so that all the interpreter overhead was bypassed.  We

### 2012-05-20 docs/paper.tex
added table centering

--- 2012-05-20 docs/paper.tex
@@
 Recently, Patrascu et al.
-[2] showed that simple tabulation hashing provides unexpectedly strong guarantees and is constant competitive with a truly random hash function for hashing with linear probing.
+showed that simple tabulation hashing provides unexpectedly strong guarantees and is constant competitive with a truly random hash function for hashing with linear probing.
 In their paper, they showed that for simple tabulation hashing with linear probing if the table size is $m = (1 + \epsilon)n$, then the expected time per operation is $O(1/\epsilon^2)$.
@@
 \begin{table}
+\centering
   \caption{Performance results on selected components of the PyBench suite.  We

### 2012-05-20 docs/paper.tex
Typos and language fixes

--- 2012-05-20 docs/paper.tex
@@
 
-In order to attempt to improve the performance of the python language as a whole we plan to attempt to improve the performance of python dictionaries.
+In order to attempt to improve the performance of the Python language as a whole we plan to attempt to improve the performance of Python dictionaries.
 
@@
 
-One of the features that we added to the python source was a conditional flag
+One of the features that we added to the Python source was a conditional flag
 that allowed us to view dictionary statistics (collisions, probes, accesses,
@@
 CPython uses some measures to optimize for this common case. Because of this
-importance we mostly limit ourselves to the study of new implementations of
-hashing the string type for the rest of the paper.
+importance we predominantly focus on the study of new implementations of
+hashes for the string type for the rest of the paper.
 
@@
 
-Linear probing is appealing because of its excellent cache behavior in comparison to the probing scheme currently utilized in Python.
-Consider the Intel Core 2 Duo processor with a 64 byte cache line.
-Each Python dictionary entry is 12 bytes so when Python’s current probing scheme probes another slot, it is not in the current cache line so it has to pull another cache line in from memory, but with linear probing, that next slot is most likely in the current cache line.
+Linear probing is appealing because of its excellent cache behavior in
+comparison to the probing scheme currently utilized in Python.  Consider the
+Intel Core 2 Duo processor with a 64 byte cache line.  Each Python dictionary
+entry is 12 bytes so when Python probes a dictionary slot, it pulls in at least
+4 more dictionary entries with it.  However, when the current probing scheme
+probes another slot, it is not in the current cache line so it has to pull
+another cache line in from memory.  With linear probing, that next slot is most likely in the current cache line.
 
-However, Patrascu et al.
+Patrascu et al.
 [2] describe simple tabulation hashing for fixed key widths.
@@
 \subsection{Implementation Details}
+We modified the source code of CPython 2.7.3 to use simple tabulation hashing
+with linear probing.  We use a fixed number of tabulation tables and cycle between them
+for variable length strings.
+
 We prepopulated random tables in memory with random bits from random.org, which
-sources randomness form atmospheric noise in order to remove the reliance on
+sources randomness from atmospheric noise in order to remove the reliance on
 pseudo-random number generators.
@@
 instructions into our hash code to tell the CPU's caching system that it should
-deprioritize storing it in cache.  We used GCC's
+deprioritize storing the tabulation tables in cache.  We used GCC's
 \begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
-to do this.  However, this only yielded about a 1\% performance improvement,
+to generate these instructions.  However, this only yielded about a 1\% performance improvement,
 negligible compared to the performance improvement of different hash schemes.
@@
 every byte separately resulted in the best performance.    We believe this is due
-to the low entropy for fewer tables and too much memory bandwidth for a higher
+to the low entropy caused by fewer tables and too much memory bandwidth used by a higher
 number of tables.  The optimal result may also be an artifact of the fact that
 we were using on the order of magnitude of thousands to millions of dictionary
-keys.  Our optimal result resulted 256 64
+keys.  Our optimal result resulted using 256 64
 bit table entries in each of 8 tables using 16KB of memory.
@@
 valuable benchmarks is to see how the new dictionary performs on them. We used
-PyBench benchmark suit for this task. We saw significant improvements of around
+PyBench benchmark suit for this task. The results can be seen in Table
+~\ref{pybenchtable}.  We saw significant improvements of around
 20\% in performance of some of the internals such as the special and normal
@@
 of the other internals such as CPython exception handler were running around
-12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measures for running the whole benchmark suite.
+12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.
 
@@
 \end{tabular}
+\label{pybenchtable}
 \end{table}
 
-PyBench showed the performance difference in each of the individual CPython internals however for a more realistic picture of the performance in real world programs more extensive benchmarking of the new implementation was needed.
-Especially since we wanted to know the frequency of the different CPython internals in real-world applications and therefore the corresponding runtime changes.
-The new implementation was tested on several standard real world programs such as Tornado web server, Django web framework, html5lib and nearly 10 other real-world programs.
+PyBench showed the performance difference in each of the individual CPython
+internals however for a more realistic picture of the performance in real world
+programs more extensive benchmarking of the new implementation was needed.
+Especially since we wanted to know the frequency of the different CPython
+internals in real-world applications and therefore the corresponding runtime
+changes. The new implementation was tested on several standard real world
+programs such as Tornado web server, Django web framework, html5lib and ten other real-world programs.
 
@@
 We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates.
-This finding followed by other benchmarks showed that the new implementation gives more consistent runtime on different iterations of a same code compared to the current CPython implementation.
-This consistency is a desired behavior.
-We suspect that this behavior is a result of the randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.
 
-As mentioned before, another set of the benchmarks that we utilized were based on different standard real-world Python applications.
+This finding as well as those by other benchmarks showed that the new
+implementation gives more consistent runtime on different iterations of a same
+code compared to the current CPython implementation. This consistency is a
+desired behavior. We suspect that this behavior is a result of the "better" randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.
+
+As mentioned before, another set of the benchmarks that we utilized was based on different standard real-world Python applications.
 We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite.
@@
   \includegraphics[width=12cm]{appsbench.pdf}
-  \caption{Percentage of overall improvements in some real-world Python applications. Nearly none of the applications performed worse. Python Pickle and html5lib faster around \%3 to \%4. }
+  \caption{Percentage of overall improvements in some real-world Python
+  applications. Nearly none of the applications performed worse. Python Pickle
+  and html5lib faster around 3\% to 4\%. }
 \end{figure}
@@
 \begin{enumerate}
-\item Hardware specific performance issues dominate at small scales: When you are attempting to work on a process that runs as fast as a hash table lookup or insertion already does, there is little room for error.
-Improving hash performance with simple tabulation is not an easy task simply because any change you make will probably end up making the hash table run slower.
-That said:
+\item Hardware specific performance issues dominate at small scales: When you
+are attempting to work on a process that runs as fast as a hash table lookup or
+insertion already does, there is little room for error.  Improving hash
+performance with simple tabulation is not an easy task simply because any
+change you make will probably end up making the hash function run slower.  That said:
 
@@
 First, we could use different dictionaries for different purposes.
-Many of the python dictionaries are small, and their keysets never change.
+Many of the Python dictionaries are small, and their keysets never change.
 No one is adding a new built-in function in the near future.
+
 Improve performance bottlenecks illustrated by benchmarks

### 2012-05-20 docs/paper.tex
fixed spell error

@@
 valuable benchmarks is to see how the new dictionary performs on them. We used
-PyBench benchmark suit for this task. The results can be seen in Table
+PyBench benchmark suite for this task. The results can be seen in Table
 ~\ref{pybenchtable}.  We saw significant improvements of around

### 2012-05-20 docs/paper.tex
Fix conclusion

@@
 
-In terms of future work, we have 2 avenues of attack.
-First, we could use different dictionaries for different purposes.
-Many of the Python dictionaries are small, and their keysets never change.
-No one is adding a new built-in function in the near future.
+In terms of future work, we have 2 avenues of attack. First, we could use
+different dictionaries for different purposes. Many of the Python dictionaries
+are small, and their keysets never change. New built in functions are not added
+during the runtime of the program and thus they could use perfect hashing.
 
-Improve performance bottlenecks illustrated by benchmarks
+Second, we could improve the performance bottlenecks illustrated by benchmarks.
+These mainly consist of the extra time and memory bandwith and latency cost for
+computing simple tabulation hashing.  If we optimize the hashing further
+through more profiling, memory layout, and instruction order optimizing, we may
+be able to achieve more real world gains.  We could also look into applying a
+modifying of the perturbation trick currently used in Python for quadratic
+probing.
 

### 2012-05-20 docs/paper.tex
fixes spelling

--- 2012-05-20 docs/paper.tex
@@
 \subsection{Theoretical Results}
-Recently, Patrascu et al.
-showed that simple tabulation hashing provides unexpectedly strong guarantees and is constant competitive with a truly random hash function for hashing with linear probing.
+Recently, P\v{a}tra\c{s}cu et al.
+\cite{mihai} showed that simple tabulation hashing provides unexpectedly strong guarantees and is constant competitive with a truly random hash function for hashing with linear probing.
 In their paper, they showed that for simple tabulation hashing with linear probing if the table size is $m = (1 + \epsilon)n$, then the expected time per operation is $O(1/\epsilon^2)$.
@@
 
+\begin{thebibliography}{10}
+
+\bibitem{mihai} Mihai P\v{a}tra\c{s}cu and Mikkel Thorup. The power of simple tabulation hashing. In \emph{Proc. of 43rd STOC}, 2011.
+
+\bibitem{python} Python Software Foundation \emph{},
+\texttt{http://python.org/} 2012.
+
+\bibitem{python} Nick Welch. Hash-table-shootout. 
+\texttt{https://github.com/mackstann/hash-table-shootout} 2012.
+
+\bibitem{python} Nicholas Piël. Benchmark of Python WSGI Servers.
+
+\texttt{http://nichol.as/benchmark-of-python-web-servers} 2010.
+
+\end{thebibliography}
 \appendix

### 2012-05-20 docs/paper.tex
fixed patrascu citation

@@
 Here we implement simple tabulation hashing with linear probing in Python
-dictionary based on the recent theoretical result of Patrascu et al. Although
+dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
 we achieved significant performance wins on synthetic benchmarks, on real
@@
 
-Patrascu et al.
-[2] describe simple tabulation hashing for fixed key widths.
+P\v{a}tra\c{s}cu et al.
+\cite{mihai} describe simple tabulation hashing for fixed key widths.
 For implementing this new hashing scheme, first we needed to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).

### 2012-05-20 docs/paper.tex
added a new page for appendix
Fixes

@@
 significant $i$ bits as the index can populate the table without a lot of
-collisions. This approach gives a better than random results in this case and is
+collisions. This approach gives better than random results in this case and is
 also simple and fast to compute.
@@
 of a hash value. We should note that the value of  perturb shift (currently 5)
-is crucial for a good probing behavior. Since it should be small enough so that
+is crucial for a good probing behavior.  It should be small enough so that
 all the significant bits have an effect on the probe sequence but at the same
@@
 
-We also experimented with indexing using every pair of bytes in the string and
-using $2^16$ tabulation table entries.
+We also experimented with indexing using every pair of bytes in the string,
+treating them as 16bit integers and
+using $2^{16}$ tabulation table entries.
 
@@
 \bibitem{python} Python Software Foundation \emph{},
-\texttt{http://python.org/} 2012.
+\texttt{http://python.org} 2012.
 
-\bibitem{python} Nick Welch. Hash-table-shootout. 
+\bibitem{hashshoot} Nick Welch. Hash-table-shootout. 
 \texttt{https://github.com/mackstann/hash-table-shootout} 2012.
 
-\bibitem{python} Nicholas Piël. Benchmark of Python WSGI Servers.
+\bibitem{wsgi} Nicholas Piël. Benchmark of Python WSGI Servers.
 
@@
 \appendix
+\newpage
+\noindent{\LARGE\bfseries Appendix}
 \section{Tornado Application}

### 2012-05-25 Objects/mihash/gmp_hash.c
misc hashes

@@ -1,0 +1,93 @@
+
+/*
+
+     USING 3.5.2
+
+ gcc -O3 -DBITS_IN=n gmp_hash.c  -lgmp
+ runs in 37.9s on Owen's machine, with 32 bits in, 32 bits out
+  (and no faster when the div and mod are removed)
+ in 34s with 16 bits in and 16 bits out.
+ in 96s    with 256 bits in and 32 bits out.
+ in 366s   with 1024 bits in and 32 bits out.
+
+ Redone so that the same amount (bits) of input data is processed as with 32 bits:
+ Real times on Owen's machine zeal (always 32 bits out):
+  32  in 48.5s 48.3s 49.8s avg 48.9
+  64  in 37.3s 37.4s 37.4s avg 37.4
+  128 in 21.8s 22.0s 22.1s avg 22.0
+  256 in 14.9s 14.7s 14.8s avg 14.8
+  512 in  12.9 12.9s 13.0s avg 12.9
+  1024 in 15.3s 15.3s 15.3 avg 15.3
+  2048 in 22.3s 22.3s 22.3savg 22.3
+
+After minor bugfix that should not have affected time by much: 
+32 -> 46.5; 46.6; 46.6
+64 -> 39.1; i
+
+ Assembly code shows it's all done via function calls (not inlined)
+ C++ version of this code was the same way
+*/
+
+
+#include <stdlib.h>
+#include <stdio.h>
+#include <gmp.h>
+//#include "/home/owen/gmp-5.0.2/gmp.h"
+
+
+#define N 1024
+#define T 1000001
+
+#ifndef BITS_IN 
+ #define BITS_IN 32
+#endif
+
+#define BITS_OUT 32
+#define K (BITS_IN + BITS_OUT)
+
+int main() {
+
+  int i,j,k;
+  mpz_t m[N+1], s[N];
+  int n1 = (N * 32)/BITS_IN;
+  gmp_randstate_t state;
+
+  unsigned long fake_dep=0;
+  
+  gmp_randinit_mt(state);
+
+  for (i=0; i < n1; ++i) {
+    mpz_init(m[i]);
+    mpz_urandomb(m[i],state,K);
+
+    mpz_init(s[i]);
+    mpz_urandomb(s[i],state,BITS_IN );
+  }
+
+  for (i=0; i < T; ++i) {
+    mpz_t res;
[…]

### 2012-05-25 Objects/mihash/hashfunctionsreal.c

@@ -1,0 +1,429 @@
+/***
+* Compile with :
+*   gcc -O3 -funroll-loops -o hashfunctionsreal hashfunctionsreal.c
+*   for some compilers (e.g., GCC 4.4) on x86-64 we need -mno-sse2 to avoid 
+*    poor performance due to bad sse instructions.
+*
+* New: For Galois Field and carryless multiplications, try this command line:
+* 
+* gcc -O3 -funroll-loops -Wall -maes -msse4 -mpclmul  -o hashfunctionsreal hashfunctionsreal.c
+*
+*
+*/
+
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/time.h>
+
+
+/* code these things in plain C to avoid surprises... */
+
+typedef unsigned int uint32;
+typedef unsigned long long uint64;
+typedef unsigned long long ticks;
+
+
+
+#if defined(LITTLEMEMORY) || defined(IPAD)
+#define N 1024* 32 // should be divisible by two!
+#else
+#define N 1024* 64 // should be divisible by two!
+#endif
+
+#ifdef IPAD
+#define TRIALS 20000
+#else
+#define TRIALS 20000
+#endif
+
+#define SHORTN  1024 // should fit in L1 cache
+#ifdef IPAD
+#define SHORTTRIALS 100000
+#else
+#define SHORTTRIALS 1000000
+#endif
+
+
+#if defined( __corei7__  )  // __amd64__ is untested
+
+// start and stop are as recommended by 
+// Gabriele Paoloni, How to Benchmark Code Execution Times on Intel® IA-32 and IA-64 Instruction Set Architectures
+// September 2010
+// http://edc.intel.com/Link.aspx?id=3954
+
+static __inline__ ticks startRDTSC (void) {
+  unsigned cycles_low, cycles_high;
+  asm volatile ("CPUID\n\t"
+		"RDTSC\n\t"
+		"mov %%edx, %0\n\t"
+		"mov %%eax, %1\n\t": "=r" (cycles_high), "=r" (cycles_low)::
+		"%rax", "%rbx", "%rcx", "%rdx");
+  return ((ticks)cycles_high << 32) | cycles_low;
+}
+
+static __inline__ ticks stopRDTSCP (void) {
+  unsigned cycles_low, cycles_high;
+/// This should work fine on most machines, if the RDTSCP thing
[…]

### 2012-05-25 Objects/mihash/owenmpfq.c

@@ -1,0 +1,83 @@
+/* get, build, install MPFQ library from INRIA */
+
+// gcc owenmpfq.c -O3 -lgmp -lmpfq_gf2n
+// old: runs in 7.69s on Owen's machine
+// new: runs in 9.6s on Owen's machine zeal
+
+#include "mpfq/mpfq_2_32.h"
+#include "stdlib.h"
+
+//hashfunctionsreal uses N=1024 T=1000000
+#define N 1024   
+#define T 1000001
+
+int main() {
+
+  int i,j,k;
+  mpfq_2_32_elt m[N], s[N];
+  // unsigned long thirtytwo = 32;
+ 
+  //int acc=0;
+  unsigned long fake_dep=0;
+
+  mpfq_2_32_field gf232;
+
+  //return foo(); // temp
+
+  mpfq_2_32_field_init(gf232);
+  //mpfq_2_32_field_specify(gf232,MPFQ_DEGREE,&thirtytwo);  // does not help
+
+
+  for (i=0; i < N; ++i) {
+    mpfq_2_32_init(gf232, &m[i]);
+    // mpfq_2_32_set_ui(gf232, m[i], (((unsigned long)rand())<<32) + rand() );
+    // mpfq_2_32_set_ui(gf232, m[i], 6UL); 
+    // the _ui stuff apparently only works for prime fields (?) from
+    // reading the perl interface source.  Big problem for me...
+
+    // however, for GF[2**32] I think we can just break the abstraction.
+
+    mpfq_2_32_random(gf232,m[i]);  // works 
+    /* m[i] = rand(); */
+
+    mpfq_2_32_init(gf232, &s[i]);
+    mpfq_2_32_random(gf232,s[i]); 
+    /* s[i] = (unsigned long) rand(); */
+  }
+
+  // mpfq_2_32_sscan(gf232, m[0], "6");  // works
+
+  
+  // printf("%d\n",(int) mpfq_2_32_get_ui(gf232,m[0])); answer in GF(2)....
+  
+  //printf("now use their print:");
+  //mpfq_2_32_print(gf232, m[0]);
+  //printf("\n");
+
+
+  for (i=0; i < T; ++i) {
+    // int res = m[0];
+    mpfq_2_32_elt res;
+    mpfq_2_32_elt temp_sum_1, temp_sum_2, temp_prod;
+    mpfq_2_32_init(gf232,&res);
[…]

### 2012-05-25 Objects/mihash/tab_test.c

@@ -1,0 +1,548 @@
+#include <stdio.h>
+
+int main() {
+    long x = 5497558133889123;
+    long table0[511] = {
+    8143882431135823606, -1891346621374036439, -7974897869020531009, 3591599807050116130, 4047365529985287222, -8443613472865579473, 7344338279563181878, 7030922722098481517, \
+    6625257078588844930, 5013971041911959535, 7772157763589206088, 4688974262828807659, 5106383932298641178, 8968934067817062883, 7447203892591461313, -1797338683085125216, \
+    -6356145513970013644, 436573673540661948, -8610637766066871293, 331326742336646684, 211871435665445795, 136339682555770032, -5589609102651440753, -1865405746700213682, \
+    -7275884913510842332, 6996806946966219294, -2448213011015614617, -4811811035920909424, 8451977770432081745, 1826031120464095115, 6737780725505362683, -6936359773898345786, \
+    2713099442511187548, 1588061177942630785, -7001466817023836904, 137441136849824044, 9105440251730191363, 4887408316096751067, -3349509464985845710, 899278167470698944, \
+    8752531768492122199, 1382778941679198865, -1030016847892215943, -2059860581788589278, -7131839436001598819, 7135083771514304225, -3565223583545873125, 4109542362632888722, \
+    4869491963873703958, -3569785846449219482, -361818513726839618, -4102339853850388770, 5892237071819630458, -6821885723362465787, -6348991131344465405, -1677239608104467874, \
+    -8471859788456931571, 3522732895755865593, 6474384878492105833, 6448904081963519143, 7257102134305341483, 4543477820029882575, -8653933761171467937, -8385539435721466483, \
+    8917066615139287152, 4929618328067067251, 2584926966900499291, -6741659515794670882, -7620398298529298744, -3448757794056163461, 8703923269773746982, -347834307098295713, \
+    -3299744126347492086, -7587784708031571258, -940384495860842223, 2076104007771062681, -8617991133823990890, -2103046968542838779, -3541338923116963783, 3020617361056160683, \
+    4808639933503121374, 7621509827469401756, -3639134745878170857, 7775162812409529747, -2767102152608212954, -6007256541541971635, 5961627282744879811, -4156278707489289506, \
+    2529529980552335540, 3416756398607484248, -7829244797726196410, 1660557360385613357, 3796244103447389367, 1420843684088151656, -8811397216151577212, 6107773594429446219, \
[…]

### 2012-05-25 docs/paper.tex

@@
 we achieved significant performance wins on synthetic benchmarks, on real
-benchmarks our wins were only on the order of 4\%.  When viewed as a whole, the
+benchmarks our wins were only on the order of 4\% due to interpreter overhead.
+Linear probing with sample tabulation hashing does improve Python performance
+and should be looked at as a replacement scheme.  As a whole, the
 extra memory bandwidth
-and latency required to compute the simple tabulation hash are too large to
-justify the dictionary speedup garnered by switching to simple tabulation hashing and linear probing.
+and latency required to compute the simple tabulation hash negates some of the
+performance improvement, but if they were able to be further mitigated, the new scheme
+should provide more than a 4\% improvement, making it even more worthwhile.
 \end{abstract}
@@
 
+In order to analyze the result of this new dictionary we measured three main
+metrics. First, we compared the performance of the simple tabulation hashing
+with Python’s current hashing scheme on random and sequential strings.
+Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs.
+Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
 \subsection{Table Number Tuning}
@@
 \subsection{Benchmarks}
-In order to analyze the result of this new dictionary we measured three main metrics.
-First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings.
-Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs.
-Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
-Because of the heavy utilization of dictionaries in Python internals one of the
-valuable benchmarks is to see how the new dictionary performs on them. We used
+Due to the heavy utilization of dictionaries in Python internals, one of the
+valuable benchmarks is to see how the new dictionary performs with them them. We used
 PyBench benchmark suite for this task. The results can be seen in Table
@@
 
-\item Simple tabulation yields slightly faster and more consistent performance: With a careful implementation simple tabulation can yield increased hash performance on linear probing work loads.
-Simple tabulation definitely leads to shorter chain lengths, however, those shorter chain lengths don’t always justify the performance cost.
+\item Simple tabulation yields slightly faster and more consistent performance:
+With a careful implementation simple tabulation can yield increased hash
+performance on linear probing work loads. Simple tabulation definitely leads to
+shorter chain lengths, however, those shorter chain lengths don’t always
+justify the performance cost in all cases.
 

### 2013-04-20 docs/paper_lncs.tex
Add lncs formatted version

--- 2012-05-25 docs/paper.tex
@@
-\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
-\author{
-        Russell Cohen \\
-            \and
-        Thomas Georgiou \\
-            \and
-        Pedram Razavi
-}
-\date{\today}
-\documentclass[11pt]{article}
-\usepackage[pdftex]{graphicx}
+\documentclass[runningheads,a4paper]{llncs}
+
+\usepackage{amssymb}
+\setcounter{tocdepth}{3}
+\usepackage{graphicx}
+\usepackage{float}
+% \usepackage{caption}
 \usepackage{setspace}
 \usepackage{fullpage}
-\usepackage{float}
+\usepackage{url}
+\usepackage{cite}
+\urldef{\mailsa}\path|{rcoh,  georgiou, prazavi}@mit.edu|
+\newcommand{\keywords}[1]{\par\addvspace\baselineskip
+\noindent\keywordname\enspace\ignorespaces#1}
 
 \begin{document}
+
+\mainmatter  % start of an individual contribution
+
+% first the title is needed
+\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
+
+% a short form should be given in case it is too long for the running head
+\titlerunning{A Reimplementation of Python’s dict}
+
+% the name(s) of the author(s) follow(s) next
+\author{Russell Cohen\and Thomas Georgiou\and Pedram Razavi}
+%
+\authorrunning{A Reimplementation of Python’s dict}
+% (feature abused for this document to repeat the title also on left hand pages)
+
+% the affiliations are given next; don't give your e-mail address
+% unless you accept that it will be published
+\institute{Massachusetts Institute of Technology\\
+77 Massachusetts Avenue  Cambridge, MA 02139, USA\\
+\mailsa\\
+\url{http://www.mit.edu}}
+
 \maketitle
-%\doublespace
 
@@
 should provide more than a 4\% improvement, making it even more worthwhile.
+\keywords{Hashing, Simple Tabulation}
 \end{abstract}
@@
 In this project we aim to analyze and improve the performance of Python's
-dictionary or dict by changing its underlying implementation.  Python is an extremely popular general purpose computing language for several reasons:
+dictionary or dict by changing its underlying implementation.  Python \cite{python} is an extremely popular general purpose computing language for several reasons:
 \begin{enumerate}
@@
 to powers of 2 so that costly modulo operations could be avoided by using bit
-masks Figure ~\ref{fig:tables} shows the performance of different numbers of
+masks Fig.~\ref{fig:tables} shows the performance of different numbers of
 tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
@@
 bit table entries in each of 8 tables using 16KB of memory.
- \begin{figure}[H]
+ \begin{figure}
    \centering
@@
 
-Figure ~\ref{fig:randstring} shows the performance for random strings.  Both
+Fig.~\ref{fig:randstring} shows the performance for random strings.  Both
 linear probing and simple tabulation result in improved performance over the
 current CPython dictionary implementation but combining both techniques results
-in over a 20\% improvement.  Figure ~\ref{fig:seqstring} shows very similar
+in over a 20\% improvement.  Fig.~\ref{fig:seqstring} shows very similar
 results for sequential strings.
+
+% \noindent%
+% \begin{minipage}{\linewidth}
+% \makebox[\linewidth]{%
+%   \includegraphics[width=4.1in]{randstring.pdf}}
+%   \captionof{figure}{Random string performance}
+%   \label{fig:randstring}
+% \end{minipage}
+
+% \noindent%
+% \begin{minipage}{\linewidth}
+% \makebox[\linewidth]{%
+%   \includegraphics[width=4.1in]{seqstring.pdf}}
+%   \captionof{figure}{Sequential string performance}
+%   \label{fig:seqstring}
+% \end{minipage}
 
@@
 
-Figure ~\ref{fig:chainlength} shows the average chain length, or dictionary
+Fig.~\ref{fig:chainlength} shows the average chain length, or dictionary
 cell probes per lookup for a run of the benchmark on sequential strings.
@@
 results in the fewest number of probes.  However, despite having the fewest
-number of probes, its real relative performance is much worse as seen in Figure
-~\ref{fig:seqstring}.  Linear probing performs faster than its average
+number of probes, its real relative performance is much worse as seen in Fig.~\ref{fig:seqstring}.  Linear probing performs faster than its average
 chain length suggests, showing the impact of looking at entries in the same
@@
 
-\begin{figure}[H]
+\begin{figure}
   \centering
@@
 We were interested to calculate if the new implementation of dictionary can handle more requests in a Python web server compared to the current CPython implementation.
-We chose Tornado web server for this task since most of the code is written in pure Python.
+We chose Tornado web server for this task since most of the code is written in pure Python \cite{wsgi}.
 An instance of Tornado server was run in a virtual machine with one assigned processor.
@@
 code compared to the current CPython implementation. This consistency is a
-desired behavior. We suspect that this behavior is a result of the "better" randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.
+desired behavior. We suspect that this behavior is a result of the ``better'' randomness of hash values in simple tabulation hashing and also the smaller average chain lengths.
 
-As mentioned before, another set of the benchmarks that we utilized was based on different standard real-world Python applications.
+As mentioned before, another set of the benchmarks \cite{hashshoot} that we utilized was based on different standard real-world Python applications.
 We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite.
@@
 
-\begin{thebibliography}{10}
+\bibliography{references}{}
+\bibliographystyle{plain}
+\end{document}
 
-\bibitem{mihai} Mihai P\v{a}tra\c{s}cu and Mikkel Thorup. The power of simple tabulation hashing. In \emph{Proc. of 43rd STOC}, 2011.
-
-\bibitem{python} Python Software Foundation \emph{},
-\texttt{http://python.org} 2012.
-
-\bibitem{hashshoot} Nick Welch. Hash-table-shootout. 
-\texttt{https://github.com/mackstann/hash-table-shootout} 2012.
-
-\bibitem{wsgi} Nicholas Piël. Benchmark of Python WSGI Servers.
-
-\texttt{http://nichol.as/benchmark-of-python-web-servers} 2010.
-
-\end{thebibliography}
 \appendix
 \newpage
-\noindent{\LARGE\bfseries Appendix}
-\section{Tornado Application}
+\section*{Appendix: Tornado Application}
 Sample Tornado 2.2.1 application used for benchmark:

### 2013-04-21 docs/paper_lncs.tex
Fix citations

@@
 \usepackage{setspace}
-\usepackage{fullpage}
+% \usepackage{fullpage}
 \usepackage{url}
@@
 77 Massachusetts Avenue  Cambridge, MA 02139, USA\\
-\mailsa\\
-\url{http://www.mit.edu}}
+\mailsa}
 
@@
 \section{Introduction}
-In this project we aim to analyze and improve the performance of Python's
+In this paper we aim to analyze and improve the performance of Python's
 dictionary or dict by changing its underlying implementation.  Python \cite{python} is an extremely popular general purpose computing language for several reasons:
@@
 In this section we give a brief overview of the key implementation details of
-dictionaries in CPython 2.7.3.  We should note that Python's dictionary supports
-several different data types as keys. However as we noted earlier, the
+dictionaries in CPython 2.7.3 \cite{python27}.  We should note that Python's dictionary supports
+several different data types as keys. However, as we noted earlier, the
 dictionaries underlying class instances and modules have only strings as keys.
@@
 
-On the other hand this hashing strategy has the tendency to fill contiguous
+On the other hand, this hashing strategy has the tendency to fill contiguous
 slots of the hash table. CPython tries to solve this behavior by utilizing a
@@
 deprioritize storing the tabulation tables in cache.  We used GCC's
-\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt}
+\begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt} \cite{prefetch}
 to generate these instructions.  However, this only yielded about a 1\% performance improvement,
@@
 
-Our primary benchmark, hash-table-shootout, is a synthetic one that called the CPython dictionary
+Our primary benchmark, hash-table-shootout \cite{hashshoot}, is a synthetic one that called the CPython dictionary
 code directly from C so that all the interpreter overhead was bypassed.  We
@@
 valuable benchmarks is to see how the new dictionary performs with them them. We used
-PyBench benchmark suite for this task. The results can be seen in Table
+PyBench benchmark suite \cite{pybench} for this task. The results can be seen in Table
 ~\ref{pybenchtable}.  We saw significant improvements of around
@@
 
-As mentioned before, another set of the benchmarks \cite{hashshoot} that we utilized was based on different standard real-world Python applications.
+As mentioned before, another set of the benchmarks \cite{bench_suite} that we utilized was based on different standard real-world Python applications.
 We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite.
@@
 \bibliographystyle{plain}
-\end{document}
 

### 2013-04-21 docs/paper_lncs.tex
Minor fix in paper

@@
 %
-\authorrunning{A Reimplementation of Python’s dict}
+\authorrunning{R. Cohen, T. Georgiou, P. Razavi}
 % (feature abused for this document to repeat the title also on left hand pages)
@@
 We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite.
-The performance of more than ten applications were recoded as shown in the figures.
+The performance of more than ten applications is indicated in Fig.~\ref{fig:suite}.
 
-\begin{figure}
+\begin{figure}[!ht]
   \centering
@@
   and html5lib faster around 3\% to 4\%. }
+  \label{fig:suite}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[!ht]
   \centering
@@
 Standard deviation of new implementation 2.0383 times smaller}
+  \label{fig:pickle}
 \end{figure}
-
-% \begin{figure}[H]
-%   \centering
-%   \includegraphics[width=160mm]{lr-10.png}
-%   \caption{}
-% \end{figure}
-
-% \begin{figure}[H]
-%   \centering
-%   \includegraphics[width=160mm]{lr-10.png}
-%   \caption{}
-% \end{figure}
 
@@
 library) and html5lib (specification parser library) we calculated an
-improvement of 3\% to 4\%. In other cases we did not calculate an speedup but
+improvement of 3\% to 4\% (Fig.~\ref{fig:pickle}). In other cases we did not calculate an speedup but
 at the same time we did not find any real-world case which the new

### 2013-04-21 docs/paper_lncs.tex
minor fixes

@@
 % first the title is needed
-\title{A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing}
+\title{A Reimplementation of Python's dict using Simple Tabulation Hashing and Linear Probing}
 
 % a short form should be given in case it is too long for the running head
-\titlerunning{A Reimplementation of Python’s dict}
+\titlerunning{A Reimplementation of Python's dict}
 
@@
 benchmarks our wins were only on the order of 4\% due to interpreter overhead.
-Linear probing with sample tabulation hashing does improve Python performance
+Linear probing with simple tabulation hashing does improve Python performance
 and should be looked at as a replacement scheme.  As a whole, the
 extra memory bandwidth
-and latency required to compute the simple tabulation hash negates some of the
-performance improvement, but if they were able to be further mitigated, the new scheme
+and latency required to compute the tabulation hash negates some of the
+performance improvements, but if they were able to be further mitigated, the new scheme
 should provide more than a 4\% improvement, making it even more worthwhile.
-\keywords{Hashing, Simple Tabulation}
+\keywords{Hash Function, Simple Tabulation, Linear Probing, Python Dictionary}
 \end{abstract}
@@
 
-Recently, with the advent of packages like numPy, sciPy and pyPy Python has began to also be used (with questionable merit) as a performance computing language due to its ease of use.
-
+Recently, with the advent of packages such as NumPy, SciPy, and PyPy Python has began to also be used (with questionable merit) as a performance computing language due to its ease of use.
 However, because Python is an interpreted language it is difficult to operate on a similar performance level to compiled languages due to the sheer overhead of the interpreter.
 
-In order to attempt to improve the performance of the Python language as a whole we plan to attempt to improve the performance of Python dictionaries.
+In order to enhance the performance of the Python language as a whole we plan to improve the performance of built-in Python dictionary, an associative array implementation which is used to store (\textit{key, value}) pairs.
 
@@
 lookups in general purpose dictionaries for a total of 116 dictionary hits. 116
-hits in a 3 line piece of code which never uses dictionaries explicitly is quite
+hits in a 3-line piece of code which never uses dictionaries explicitly is quite
 significant.  We explain this extensive dictionary usage below.
@@
 In this section we give a brief overview of the key implementation details of
-dictionaries in CPython 2.7.3 \cite{python27}.  We should note that Python's dictionary supports
+dictionaries in CPython 2.7.3 \cite{python27}, the default and most-widely used implementation
+of Python which is written in C.  We should note that Python's dictionary supports
 several different data types as keys. However, as we noted earlier, the
@@
 value for sequential strings differ only in the least significant bits, for
-instance: hash("6.851") = 7824665118634166871 and hash("6.852") =
+instance: hash(``6.851'') = 7824665118634166871 and hash(``6.852'') =
 7824665118634166868. This behavior gives good results for sequential strings
@@
 On the other hand, this hashing strategy has the tendency to fill contiguous
-slots of the hash table. CPython tries to solve this behavior by utilizing a
-good collision resolution strategy. The collision resolution strategy currently
-used is open addressing with a custom non-linear probing scheme. In this scheme
+slots of the hash table. CPython tries to remedy this issue by utilizing a
+reliable collision resolution strategy. The collision resolution strategy currently
+used is open addressing combined with a custom non-linear probing mechanism. In this scheme
 CPython uses quadratic probing based on this function: $j = (5j + 1) \bmod 2^i$
-where $0<j<2^i-1$. This recurrence alone is not enough for a better behavior
-since again it behaves the same as linear-probing for some keys since it scans
-the table entries in a fixed order. To overcome this issue CPython also uses the
-most significant bits of the hash value with the following code snippet:
+where $0<j<2^i-1$. This recurrence alone is not enough to achieve a faster runtime. 
+For a set of keys, this strategy behaves the same as linear-probing because it scans
+the table entries in a fixed order. To overcome this problem, CPython also uses the
+most significant bits of the hash value by using the following code snippet:
 
@@
 perturb = hash;
-while (<slot is full> && <item in slot does not equal the key>) {
-       slot = (5*slot) + 1 + perturb;
-       perturb >>= 5;
+while (<slot is full> AND <item in slot does not equal the key>) {
+       slot = (5 * slot) + 1 + perturb;
+       perturb = (perturb >> 5); // Shifted right by 5 bits
 }
@@
 all the significant bits have an effect on the probe sequence but at the same
-time it should be large enough so that in really bad cases the high-order hash
+time, it should be large enough so that in collision-prone cases the high-order hash
 bits can affect the earlier iterations.
@@
 sparse and put a bound on the number of collisions that can happen. CPython
-resizes the size of table whenever the load factor invariant gets violated. It
+resizes a table whenever the load factor invariant gets violated. It
 quadruples the size when there are less than 50,000 keys in the table and
@@
 Overall, the design of CPython dictionary is simple and is optimized for more
-common cases such as sequential strings. Since it utilizes quadratic probing
+common cases such as retrieval of sequential strings. Since it utilizes quadratic probing
 with perturbation, even if a data type does not provide a good hash function,
-the table still gets populated with with short chain lengths.  On the other hand
-there are some behaviors which can be improved. For instance there is a bad
-cache locality because for probing a slot the table might need to jump to the
-other slots which are not immediate neighbors and not necessarily in cache.
+the table still gets populated with short chain lengths. However, there is still more room for 
+improvements and optimizations. For instance, the current approach incurs suboptimal cache locality because probing a single
+slot might require jumping to the other slots which are not its immediate 
+neighbors and consequently, not necessarily stored in the cache.
 
@@
 P\v{a}tra\c{s}cu et al.
-\cite{mihai} describe simple tabulation hashing for fixed key widths.
+\cite{mihai} describe simple tabulation hashing for fixed key lengths.
 For implementing this new hashing scheme, first we needed to figure out a variant of simple tabulation hashing for variable length strings that does not require a table for every possible index (we could potentially cycle through a fixed set of tables).
-Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing so we will have to be careful about how to implement this.
+Performing table lookups is potentially an expensive operation with lots of cache hits mitigating the better cache behavior of linear probing, therefore we have to be careful about how it will be implemented.
 In the following section we explain the implementation details of our approach to these problems.
@@
 while (--len >= 0) {
-        register long index = (*p++) | ((len&TABLE_MASK)<<8);
+  register long index = (*p++) | ((len & TABLE_MASK) << 8);
   x = x ^ RAND_TABLE_NAME[index];
@@
 finds the item it is looking for or an empty cell.
+
 \subsubsection{Optimizations}
@@
 metrics. First, we compared the performance of the simple tabulation hashing
-with Python’s current hashing scheme on random and sequential strings.
+with Python's current hashing scheme on random and sequential strings.
 Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs.
@@
 we were using on the order of magnitude of thousands to millions of dictionary
-keys.  Our optimal result resulted using 256 64
-bit table entries in each of 8 tables using 16KB of memory.
+keys.  Our optimal result resulted using 256 64-bit table entries in each of 8 tables using 16KB of memory.
  \begin{figure}
@@
 Due to the heavy utilization of dictionaries in Python internals, one of the
-valuable benchmarks is to see how the new dictionary performs with them them. We used
+valuable benchmarks is to see how the new dictionary performs with them. We used
 PyBench benchmark suite \cite{pybench} for this task. The results can be seen in Table
@@
 PyBench showed the performance difference in each of the individual CPython
-internals however for a more realistic picture of the performance in real world
+internals; however, for a more realistic picture of the performance in real world
 programs more extensive benchmarking of the new implementation was needed.
@@
 Considering the variance of request rates on different iterations of the test, we could not find any significant improvement of the response rate of the new implementation.
-We found that on the average performance was similar however the new implementation showed a smaller standard deviation in the response rates.
+We found that on the average performance was similar, however, the new implementation showed a smaller standard deviation in the response rates.
 
@@
 
-As mentioned before, another set of the benchmarks \cite{bench_suite} that we utilized was based on different standard real-world Python applications.
+As mentioned before, another set of the benchmarks \cite{bench_suite} that we utilized was based on standard real-world Python applications.
 We ran and timed the performance of the new implementation and CPython 2.7.3 side by side using the Python-Mirrors suite.
@@
 performance on linear probing work loads. Simple tabulation definitely leads to
-shorter chain lengths, however, those shorter chain lengths don’t always
+shorter chain lengths, however, those shorter chain lengths do not always
 justify the performance cost in all cases.

### 2013-04-22 docs/paper_lncs.tex
Add in machine information
New abstract

@@
 \begin{abstract}
-Here we implement simple tabulation hashing with linear probing in Python
-dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
-we achieved significant performance wins on synthetic benchmarks, on real
-benchmarks our wins were only on the order of 4\% due to interpreter overhead.
-Linear probing with simple tabulation hashing does improve Python performance
+We reimplement the Python dictionary using simple tabulation hashing with linear probing based on the recent theoretical result of P\v{a}tra\c{s}cu et al.
+This hashing scheme provides expected constant time per operation in theory and we analyze its performance in practice compared to existing schemes.
+Our new implementation of Python's \textbf{dict} achieves significant performance wins on synthetic benchmarks.  
+Overall Python performance on real world benchmarks improves by about 4\%.
+Using linear probing with simple tabulation hashing does improve Python performance
 and should be looked at as a replacement scheme.  As a whole, the
@@
 Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
+We ran the following on an Intel Core 2 Duo P8400 processor (3M cache, 2.26GHz) w/ 4GB of 1066MHz DDR3 RAM.
 \subsection{Table Number Tuning}

### 2013-04-22 docs/paper_lncs.tex
New conclusion

@@
 with linear probing.  We use a fixed number of tabulation tables and cycle between them
-for variable length strings.
+for variable length strings. This does not provide the same level of theoretical guarantees for strings with length longer than the number of tables we have but empirically provides good results.
 
@@
 treating them as 16bit integers and
-using $2^{16}$ tabulation table entries.
+using $2^{16}$ tabulation table entries but results were worse due to increased memory thrashing.
 
@@
 over twice as many level 1 cache misses with tabulation hashing than without.
-Hypothesizing that the tabulation tables were evicting other, more relevant
-data from the level 1 cache.  To fix this, we inserted x86 memory prefetch
+Hypothesizing that the cache misses were due to the tabulation tables not being in processor cache (16KB tables, 32KB L1 data, 3MB L2), we inserted x86 memory prefetch
 instructions into our hash code to tell the CPU's caching system that it should
-deprioritize storing the tabulation tables in cache.  We used GCC's
+prioritize storing the tabulation tables in cache.  We used GCC's
 \begin{tt}\_\_builtin\_prefetch(void \*addr, char rw, char locality)\end{tt} \cite{prefetch}
-to generate these instructions.  However, this only yielded about a 1\% performance improvement,
+to generate these instructions.  However, this only yielded about a 1\% performance improvement in synthetic benchmarks,
 negligible compared to the performance improvement of different hash schemes.
@@
 \section{Conclusion and future work}
-Throughout the course of our work we came to the following conclusions:
-\begin{enumerate}
-\item Hardware specific performance issues dominate at small scales: When you
-are attempting to work on a process that runs as fast as a hash table lookup or
-insertion already does, there is little room for error.  Improving hash
-performance with simple tabulation is not an easy task simply because any
-change you make will probably end up making the hash function run slower.  That said:
+%Throughout the course of our work we came to the following conclusions:
+%\begin{enumerate}
+%\item Hardware specific performance issues dominate at small scales: When you
+%are attempting to work on a process that runs as fast as a hash table lookup or
+%insertion already does, there is little room for error.  Improving hash
+%performance with simple tabulation is not an easy task simply because any
+%change you make will probably end up making the hash function run slower.  That said:
 
-\item Simple tabulation yields slightly faster and more consistent performance:
-With a careful implementation simple tabulation can yield increased hash
-performance on linear probing work loads. Simple tabulation definitely leads to
-shorter chain lengths, however, those shorter chain lengths do not always
-justify the performance cost in all cases.
+Simple tabulation yields faster and more consistent hash performance than Python's current implementation  despite the increased number of memory accesses needed to compute the simple tabulation hash.
+Our results show that performance improvements of up to 19\% are realizable in certain benchmarks with heavy lookup workloads such as class attribute lookups.
+Overall improvement real python code performance improved by about 4\%.
+%Simple tabulation definitely leads to shorter chain lengths, however, those shorter chain lengths do not %always justify the performance cost in all cases.
 
-\item Memory latency presents serious but workable issues.
-Doing a look up in the simple-tabulation table \emph{requires} a random memory access, exactly what we were trying to avoid.
-With careful memory management we can maintain good performance, but a solution requires careful thought about cache sizes and special purpose compiler directives.
-
-\end{enumerate}
+%Memory latency presents the most serious perofmrance issue.
+Doing a look up in the simple-tabulation %table \emph{requires} a random memory access, exactly what we were trying to avoid.
+With careful memory %management we can maintain good performance, but a solution requires careful thought about cache sizes %and special purpose compiler directives.
 
@@
 are small, and their keysets never change. New built in functions are not added
-during the runtime of the program and thus they could use perfect hashing.
+during the runtime of the program and thus they could use perfect hashing to improve lookups.
 
@@
 Given the strong performance of simple tabulation hashing and the only downside
-being the cost of memory lookups, processor manufacturers should look into
-adding random tables into hardware as a ROM.  We found that for what we found
-optimal for Python, we only needed 16KB of randomness.  This would be a trivial
+being the cost of memory lookups hampering performance, processor manufacturers should look into
+adding random tables into hardware as a ROM.  We found that we only needed 16KB of randomness to acheive our 4\% overall performance gains for Python. This would be a trivial
 cost in ROM given that processors have MBs of higher level caches nowadays.
+An x86 instruction set extension providing access to random tables would benefit numerous applications that use dictionaries as a core data structures.
 
+Overall, we suggest that the Python community should consider simple tabulation hashing with linear probing as a better implementation of Python's \textbf{dict} due to realizable performance improvements of at least 4\% and lessened performance variation.
+In addition, if the tables are randomized at interpreter startup, the potential for malicious attacks against dictionary performance could be remedied.
 \bibliography{references}{}
@@
 
-\appendix
-\newpage
-\section*{Appendix: Tornado Application}
-Sample Tornado 2.2.1 application used for benchmark:
- \begin{verbatim}
-import tornado.httpserver
-import tornado.ioloop
-import tornado.wsgi
-
-def application(environ, start_response):
-    status = '200 OK'
-    output = 'Hello World!' 
-    response_headers = [('Content-type', 'text/plain'),
-        ('Content-Length', str(len(output)))]
-    start_response(status, response_headers)
-    return [output]
-def main():
-    container = tornado.wsgi.WSGIContainer(application)
-    http_server = tornado.httpserver.HTTPServer(container)
-    http_server.listen(8000)
-    tornado.ioloop.IOLoop.instance().start()
-if __name__ == "__main__":
-    main()
- \end{verbatim}
-\appendix
-
 \end{document}
